---
title: "Linear Regression"
subtitle: "A linear regression plot predicting athletes' heights based on their weights."
format: html
engine: knitr
---

Recreates [Linear
Regression](https://idl.uw.edu/mosaic/examples/linear-regression.html)
from the mosaic/vgplot example gallery. Regression computation is
performed in the database. The area around a regression line shows a
95% confidence interval. Select a region to view regression results for
a data subset.

The brushed selection (`$query`, created via `intervalXY(as = query)`)
is used only within this one plot, so -- unlike "Sorted Bars" -- it
doesn't need an explicit `vg_params()` declaration; the selection is
created the moment it's targeted by an `as =`.

```{r setup}
library(vgplotr)
```

```{r}
query <- param(query)

spec <-
  vg_create() |>
  vg_meta(
    title = "Linear Regression",
    description = paste(
      "A linear regression plot predicting athletes' heights based on their weights.",
      "Regression computation is performed in the database. The area around a regression",
      "line shows a 95% confidence interval. Select a region to view regression results",
      "for a data subset."
    )
  ) |>
  vg_data(name = "athletes", file = vg_data_url("athletes.parquet")) |>
  vg_mark_dot(data_from = "athletes", x = ~weight, y = ~height, fill = ~sex, r = 2, opacity = 0.05) |>
  vg_mark_regression_y(data_from = "athletes", filter_by = query, x = ~weight, y = ~height, stroke = ~sex) |>
  vg_interval_xy(as = query, brush = list(fillOpacity = 0, stroke = "currentColor")) |>
  vg_plot(xyDomain = "Fixed", colorDomain = "Fixed")

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/linear-regression.html)
exactly, aside from `data.athletes.file` (`vg_data_url()` vs. a relative
path) and object key order.
:::
