---
title: "Line Multi-Series"
subtitle: "Unemployment rate of various U.S. metro divisions, with hover-highlighting of the nearest series."
format: html
engine: knitr
---

Recreates [Line
Multi-Series](https://idl.uw.edu/mosaic/examples/line-multi-series.html)
from the mosaic/vgplot example gallery. On hover, the closest data point
to the pointer and its associated series is highlighted. Highlighting
of series is performed using `nearestX` and `highlight` interactors.
Point and text annotations instead use the mark `select` filter option.

```{r setup}
library(vgplotr)
```

```{r}
curr <- param(curr)

spec <-
  vg_create() |>
  vg_meta(
    title = "Line Multi-Series",
    description = paste(
      "This line chart shows the unemployment rate of various U.S. metro divisions from 2000",
      "through 2013. On hover, the closest data point to the pointer and its associated",
      "series is highlighted. Highlighting of series is performed using `nearestX` and",
      "`highlight` interactors. Point and text annotations instead use the mark `select`",
      "filter option."
    ),
    credit = paste(
      "Adapted from a [D3 example](https://observablehq.com/@d3/multi-line-chart/2). Data",
      "from the [Bureau of Labor Statistics](https://www.bls.gov/)."
    )
  ) |>
  vg_data(name = "bls_unemp", file = vg_data_url("bls-metro-unemployment.parquet")) |>
  vg_mark_rule_y(data_from = c(0)) |>
  vg_mark_line_y(
    data_from = "bls_unemp", data_optimize = FALSE,
    x = ~date, y = ~unemployment, z = ~division,
    stroke = "steelblue", stroke_opacity = 0.9, curve = "monotone-x"
  ) |>
  vg_nearest_x(channels = list("z"), as = curr) |>
  vg_highlight(by = curr) |>
  vg_mark_dot(
    data_from = "bls_unemp", x = ~date, y = ~unemployment, z = ~division,
    r = 2, fill = "currentColor", select = "nearestX"
  ) |>
  vg_mark_text(
    data_from = "bls_unemp", x = ~date, y = ~unemployment,
    text = ~division, fill = "currentColor", dy = -8, select = "nearestX"
  ) |>
  vg_plot(
    margin_left = 24, x_label = NULL, x_ticks = 10, y_label = "Unemployment (%)", y_grid = TRUE,
    style = "overflow: visible;", width = 680
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/line-multi-series.html)
exactly, aside from `data.bls_unemp.file` (`vg_data_url()` vs. a
relative path) and object key order.
:::
