---
title: "Normalized Stock Prices"
subtitle: "Hover over the chart to normalize stock prices for the percentage return on a given day."
format: html
engine: knitr
---

Recreates [Normalized Stock
Prices](https://idl.uw.edu/mosaic/examples/normalize.html) from the
mosaic/vgplot example gallery. A `nearestX` interactor selects the
nearest date, and parameterized expressions reactively update in
response.

The `point` param is declared as a mosaic `ParamDate` (`list(date =
"2013-05-13")`, mosaic-spec's `{"date": "..."}` shape), and used both
directly as a mark encoding (`x = point`, with no `~` formula needed,
since a `param()` reference is already a self-contained value) and
interpolated inside `sql()` expressions.

```{r setup}
library(vgplotr)
```

```{r}
point <- param(point)

spec <-
  vg_create() |>
  vg_meta(
    title = "Normalized Stock Prices",
    description = paste(
      "What is the return on investment for different days? Hover over the chart to",
      "normalize the stock prices for the percentage return on a given day. A `nearestX`",
      "interactor selects the nearest date, and parameterized expressions reactively update",
      "in response."
    )
  ) |>
  vg_data(name = "stocks", file = vg_data_url("stocks.parquet")) |>
  vg_data(
    name = "labels",
    query = "SELECT MAX(Date) as Date, ARGMAX(Close, Date) AS Close, Symbol FROM stocks GROUP BY Symbol"
  ) |>
  vg_params(point = list(date = "2013-05-13")) |>
  vg_mark_rule_x(x = point) |>
  vg_mark_text_x(x = point, text = point, frame_anchor = "top", line_anchor = "bottom", dy = -7) |>
  vg_mark_text(
    data_from = "labels", x = ~Date,
    y = ~ sql("Close / (SELECT max(Close) FROM stocks WHERE Symbol = source.Symbol AND Date = ", point, ")"),
    dx = 2, text = ~Symbol, fill = ~Symbol, text_anchor = "start"
  ) |>
  vg_mark_line_y(
    data_from = "stocks", x = ~Date,
    y = ~ sql("Close / (SELECT max(Close) FROM stocks WHERE Symbol = source.Symbol AND Date = ", point, ")"),
    stroke = ~Symbol
  ) |>
  vg_nearest_x(as = point) |>
  vg_plot(
    y_scale = "log", y_domain = c(0.2, 6), y_grid = TRUE, x_label = NULL, y_label = NULL,
    y_tick_format = "%", width = 680, height = 400, margin_right = 35
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/normalize.html) exactly, aside
from `data.stocks.file` (`vg_data_url()` vs. a relative path) and
object key order.
:::
