---
title: "Wind Map"
subtitle: "Vector marks on a grid show wind direction and intensity."
format: html
engine: knitr
---

Recreates [Wind Map](https://idl.uw.edu/mosaic/examples/wind-map.html)
from the mosaic/vgplot example gallery (adapted from an [Observable
Plot example](https://observablehq.com/@observablehq/plot-wind-map)).
Expressions for `rotate`, `length`, and `stroke` values are evaluated
in the database. Both the legend and map support interactive
selections to highlight values.

```{r setup}
library(vgplotr)
```

```{r}
selected <- param(selected)
length_p <- param(length)

spec <-
  vg_create() |>
  vg_meta(
    title = "Wind Map",
    description = paste(
      "`vector` marks on a grid show both direction and intensity—here, the speed of",
      "winds. Expressions for `rotate`, `length`, and `stroke` values are evaluated in the",
      "database. Both the legend and map support interactive selections to highlight values."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-wind-map)."
  ) |>
  vg_data(name = "wind", file = vg_data_url("wind.parquet"), select = c("*", "row_number() over () as id")) |>
  vg_params(selected = list(select = "union"), length = 2) |>
  vg_vconcat(
    vg_legend_color(for_plot = "wind-map", label = "Speed (m/s)", as = selected),
    vg_mark_vector(
      data_from = "wind", x = ~longitude, y = ~latitude,
      rotate = ~ sql("degrees(atan2(u, v))"),
      length = ~ sql(length_p, " * sqrt(u * u + v * v)"),
      stroke = ~ sql("sqrt(u * u + v * v)"),
      channels = list(id = "id")
    ) |>
      vg_region(as = selected, channels = list("id")) |>
      vg_highlight(by = selected) |>
      vg_plot(
        name = "wind-map", length_scale = "identity", color_zero = TRUE,
        inset = 10, aspect_ratio = 1, width = 680
      ),
    vg_slider(min = 1, max = 7, step = 0.1, as = length_p, label = "Vector Length")
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/wind-map.html) exactly, aside
from `data.wind.file` (`vg_data_url()` vs. a relative path) and object
key order.
:::
