---
title: "WNBA Shot Chart"
subtitle: "Every field goal attempt in the 2023 WNBA regular season, grouped into hexagonal bins."
format: html
engine: knitr
---

Recreates [WNBA Shot
Chart](https://idl.uw.edu/mosaic/examples/wnba-shots.html) from the
mosaic/vgplot example gallery. Shots are grouped into hexagonal bins,
with color indicating shot potency (average score) and size indicating
the total count of shots per location. The menu filters isolate shots
by team or athlete.

The two menus share one `crossfilter` Selection (`filter`): the
athlete menu both filters by it (`filter_by = filter_p`, respecting the
team menu's choice) and contributes its own clause to it (`as =
filter_p`), so team and athlete filters combine.

```{r setup}
library(vgplotr)
```

```{r}
filter_p <- param(filter)
bin_width <- param(binWidth)

spec <-
  vg_create() |>
  vg_meta(
    title = "WNBA Shot Chart",
    description = paste(
      "Every field goal attempt in the 2023 WNBA regular season. Shots are grouped into",
      "hexagonal bins, with color indicating shot potency (average score) and size",
      "indicating the total count of shots per location. The menu filters isolate shots by",
      "team or athlete."
    ),
    credit = paste(
      "Data from [Wehoop](https://wehoop.sportsdataverse.org/). Design inspired by",
      "[Kirk Goldsberry](https://en.wikipedia.org/wiki/Kirk_Goldsberry) and a",
      "[UW CSE 512](https://courses.cs.washington.edu/courses/cse512/24sp/) project by",
      "Mackenzie Pitts and Madeline Brown."
    )
  ) |>
  vg_data(
    name = "shots", file = vg_data_url("wnba-shots-2023.parquet"),
    where = "NOT starts_with(type, 'Free Throw') AND season_type = 2"
  ) |>
  vg_data(name = "court", file = vg_data_url("wnba-half-court.parquet")) |>
  vg_params(filter = list(select = "crossfilter"), binWidth = 18) |>
  vg_vconcat(
    vg_hconcat(
      vg_menu(from = "shots", column = "team_name", as = filter_p, label = "Team"),
      vg_menu(from = "shots", column = "athlete_name", filter_by = filter_p, as = filter_p, label = "Athlete")
    ),
    vg_vspace(5),
    vg_mark_frame(stroke_opacity = 0.5) |>
      vg_mark_hexgrid(bin_width = bin_width, stroke_opacity = 0.05) |>
      vg_mark_hexbin(
        data_from = "shots", filter_by = filter_p, bin_width = bin_width,
        x = ~x_position, y = ~y_position, fill = ~ vg_avg(score_value), r = ~ vg_count(),
        tip = list(format = list(x = FALSE, y = FALSE))
      ) |>
      vg_mark_line(data_from = "court", stroke_linecap = "butt", stroke_opacity = 0.5, x = ~x, y = ~y, z = ~z) |>
      vg_plot(
        name = "shot-chart", xAxis = NULL, yAxis = NULL, margin = 5,
        xDomain = c(0, 50), yDomain = c(0, 40), colorDomain = "Fixed", colorScheme = "YlOrRd",
        colorScale = "linear", colorLabel = "Avg. Shot Value", rScale = "log", rRange = c(3, 9),
        rLabel = "Shot Count", aspectRatio = 1, width = 510
      ),
    vg_legend_color(for_plot = "shot-chart")
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/wnba-shots.html) exactly,
aside from `data.shots.file`/`data.court.file` (`vg_data_url()` vs. a
relative path) and object key order.
:::
