---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# {vgplotr}: An R interface to Mosaic vgplot

## What is Mosaic?

Mosaic is a framework for linking data visualizations, tables, input widgets, and other data-driven components, while leveraging a database for scalable processing. With Mosaic, you can interactively visualize and explore millions and even billions of data points.

A key idea is that interface components – Mosaic clients – publish their data needs as queries that are managed by a central coordinator. The coordinator may further optimize queries before issuing them to a backing data source such as DuckDB.

Mosaic/vgplot is a grammar of interactive graphics in 
which graphical marks are Mosaic clients.

As the name suggests, vgplot combines concepts from existing tools such as Vega-Lite, ggplot2, and Observable Plot. Like Vega-Lite, vgplot supports rich interactions and declarative specification either using an API or standalone JSON/YAML specs (via the mosaic-spec package). However, because vgplot is based on Mosaic, it interoperates with other Mosaic clients, such as the included Mosaic Inputs. vgplot calls Observable Plot to render SVG output.

## Bringing Mosaic to R

The goal of {vgplotr} is to bring Mosaic functionality to R
in a way that feels familiar to R users.

```{r}
library(vgplotr)
library(mosaicData)
Births2015 |>
  vg_mark_line(x = ~date, y = ~births, stroke = ~wday) |>
  vg_meta(title = "US Births in 2015") |>
  vg_render()
```

Here is an example that shows some of the interactive
capabilities of {vgplotr}:

```{r}
#| label: stocks
#| code-fold: true
library(vgplotr)

point <- param(point)

vg_create() |>
  vg_meta(
    title = "Normalized Stock Prices",
    description = paste(
      "What is the return on investment for different days? Hover over the chart to",
      "normalize the stock prices for the percentage return on a given day. A `nearestX`",
      "interactor selects the nearest date, and parameterized expressions reactively update",
      "in response."
    )
  ) |>
  vg_data(name = "stocks", file = vg_data_url("stocks.parquet")) |>
  vg_data(
    name = "labels",
    query = "SELECT MAX(Date) as Date, ARGMAX(Close, Date) AS Close, Symbol FROM stocks GROUP BY Symbol"
  ) |>
  vg_params(point = list(date = "2013-05-13")) |>
  vg_mark_rule_x(x = point) |>
  vg_mark_text_x(
    x = point,
    text = point,
    frame_anchor = "top",
    line_anchor = "bottom",
    dy = -7
  ) |>
  vg_mark_text(
    data_from = "labels",
    x = ~Date,
    y = ~ sql(
      "Close / (SELECT max(Close) FROM stocks WHERE Symbol = source.Symbol AND Date = ",
      point,
      ")"
    ),
    dx = 2,
    text = ~Symbol,
    fill = ~Symbol,
    text_anchor = "start"
  ) |>
  vg_mark_line_y(
    data_from = "stocks",
    x = ~Date,
    y = ~ sql(
      "Close / (SELECT max(Close) FROM stocks WHERE Symbol = source.Symbol AND Date = ",
      point,
      ")"
    ),
    stroke = ~Symbol
  ) |>
  vg_nearest_x(as = point) |>
  vg_plot(
    y_scale = "log",
    y_domain = c(0.2, 6),
    y_grid = TRUE,
    x_label = NULL,
    y_label = NULL,
    y_tick_format = "%",
    width = 680,
    height = 400,
    margin_right = 35
  ) |>
  vg_render()
```

<!-- badges: start -->
<!-- badges: end -->


## Installation

Once the GitHub repo has been made public, you can install the 
development version of vgplotr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("rpruim/vgplotr")
```

