---
title: "Custom JavaScript with vgplotr"
vignette: >
  %\VignetteIndexEntry{Custom JavaScript with vgplotr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
toc: true
engine: knitr
---

```{r}
#| label: setup
#| include: false
library(vgplotr)
set.seed(12345)
# Reference DuckDB-Wasm from the CDN in this vignette's *built* output,
# regardless of whether the machine building it happens
# to have a local cache (vg_cache_duckdb()).  See ?vg_widget's `use_cache`.
options(vgplotr.use_cache = FALSE)
```

Mosaic specs are data: they can name a column or hold a literal, but they can't
hold code. Most graphics never need to. This article is for the times when
one does, and it assumes you're comfortable reading a little JavaScript. It
covers two functions, both of which run in the browser, with the page's own
privileges. As with any JavaScript you put in a page, only use code you wrote or
trust.

- `js()` supplies a *value*, most usefully a function, for an option.
- `vg_on_render()` runs a function once the graphic has rendered, with access to
  its params, selections and database queries.

New to vgplotr? Start with `vignette("getting-started")`.

## Function-valued options: `js()`

`js()` marks a string of JavaScript as a *value* rather than text, so it reaches
the plot as a function instead of being read as a column name. Observable Plot
accepts a function for many options, called with the row's index as its second
argument (mosaic keeps data in columns, so the first argument is `undefined`).
Here it dims every other point:

```{r}
#| label: js-opacity

alternating <- data.frame(x = 1:20, y = cumsum(rnorm(20)))

alternating |>
  vg_mark_dot(
    x = ~x, y = ~y, r = 6, fill = "steelblue",
    opacity = js("(d, i) => i % 2 ? 1 : 0.25")
  ) |>
  vg_render()
```

Because the function only sees the row's position, `js()` suits constants and
values computed from the index. To draw from a column's values, map the column
itself (`opacity = ~col`) or use `sql()`. See `?js`.

## Custom behavior: `vg_on_render()`

Some things a spec can't express at all: a *client* that runs its own query and
reacts to a selection, or a *clause* you publish into a selection from your own
controls. `vg_on_render()` takes a JavaScript function that vgplotr calls once
the graphic has rendered. It receives one object holding everything it needs:

| Property | What it is |
|---|---|
| `element` | the graphic's root DOM element; append your own elements to it |
| `params` | a `Map` of every param and selection, including ones that are only referred to, like an interactor's `as = param(brush)` |
| `coordinator` | the graphic's mosaic coordinator, which runs its queries |
| `vg` | the vgplot API bound to that coordinator, including the SQL builders (`vg.Query`, `vg.sql`, `vg.count()`, ...) |
| `mosaic` | the `mosaic-core` module: `MosaicClient`, `Selection`, and the clause helpers such as `mosaic.clauseInterval()` |
| `makeClient(options)` | mosaic's `makeClient()`, already connected to this graphic |

Use this `makeClient()` rather than `mosaic.makeClient()`, which connects to a
different, global coordinator.

### An example: a live count and a toggle

Below, the graphic adds a label with a custom client that counts the rows
inside the brush, plus a button that restricts the selection to `x` between 40 and 60
and takes the restriction off again. The client re-runs its query whenever the
selection changes, whether the change comes from the brush or from the button,
and the two combine: with both, only points in the overlap are counted. (The
plot below the brush shows only the selected points, so you can see the effect.)

```{r}
#| label: on-render

uniform_points <- data.frame(x = runif(200, 0, 100), y = runif(200, 0, 100))

vg_create() |>
  vg_data("uniform_points", uniform_points) |>
  vg_params(brush = vg_selection("crossfilter")) |>
  vg_vconcat(
    vg_mark_dot(
      x = ~x, y = ~y, r = 3, fill = "gray", data_from = "uniform_points",
      width = 500, height = 160
    ) |>
      vg_interval_x(as = param(brush)),
    vg_mark_dot(
      x = ~x, y = ~y, r = 3, fill = "steelblue", data_from = "uniform_points",
      filter_by = param(brush), width = 500, height = 160
    )
  ) |>
  vg_on_render(js("({ element, params, makeClient, vg, mosaic }) => {
    const brush = params.get('brush');
    const panel = element.appendChild(document.createElement('div'));
    const label = panel.appendChild(document.createElement('span'));
    const button = panel.appendChild(document.createElement('button'));

    // The restriction is on while the button's own clause is in the selection.
    // Reading that back, instead of keeping a separate flag, means the button
    // is right however the selection changes.
    const restricted = () => brush.clauses.some(clause => clause.source === button);
    const sync = () => {
      button.textContent = restricted() ? 'Remove [40, 60] restriction' : 'Restrict to [40, 60]';
    };
    brush.addEventListener('value', sync);
    sync();

    // A clause with no interval takes the button's clause back out.
    button.onclick = () => brush.update(
      mosaic.clauseInterval('x', restricted() ? null : [40, 60], { source: button })
    );

    makeClient({
      selection: brush,
      query: filter => vg.Query.from('uniform_points').select({ n: vg.count() }).where(filter),
      queryResult: data => { label.textContent = [...data][0].n + ' of 200 points selected  '; }
    });
  }")) |>
  vg_render()
```

A few things to notice:

- The count comes from a *client*: `makeClient()` is given the selection to
  listen to (`selection: brush`) and a `query` function that turns the
  selection's current filter into SQL. Whenever the selection changes, mosaic
  calls `query` again and hands the result to `queryResult`.
- The button publishes a *clause*: `mosaic.clauseInterval()` builds one for a
  column and an interval, tagged with the `source` that made it. Publishing
  another clause from the same source replaces the earlier one, and a clause with
  no interval (`null`) removes it, which is how the brush clears itself too.
- The table is named `uniform_points`, rather than something generic, because every
  graphic on a page shares one database. Two graphics on one page that use the
  same table name for different data overwrite each other.

The function may be `async`, and you can add several hooks with more calls to
`vg_on_render()`; they run in order. If a hook throws, the error is shown under
the graphic.

## What to know

- The code runs once per rendering, in the browser, and can only see what the
  context object gives it.
- Hooks are not part of mosaic-spec, so `to_json()` and `to_yaml()` leave them
  out, with a warning. Likewise they apply to a spec built with `vg_create()`,
  not to a JSON or YAML string given to `vg_widget()`.
- Hooks are for live output. In a non-HTML output, where `vg_render()` falls
  back to a static snapshot, there is nothing to click.
