---
title: "{vgplotr} Design Principles"
vignette: >
  %\VignetteIndexEntry{{vgplotr} Design Principles}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


```{r}
#| label: setup
#| include: false
library(tibble)
library(dplyr)
library(gt)
library(vgplotr)
library(palmerpenguins)
```

Part of learning a new graphics system is learning the 
principles that were used in its design.  This
vignette lays out some of the important vocabulary,
concepts, and design principles used to create {vgplotr},
many of which are inherited either from mosaic or
from the grammar of graphics more generally.

# What is a graphic?  What is a grammar of graphics?

A **graphic** is a visual^[Perhaps *sensory* would be 
better than visual, since once could make tactile
representations of data as well. But since {vgplotr}
is concerned with *visual* representations, we will
stick with visual here.] 
display of data.  Leland Wilkerson proposed a 
**grammar of graphics** that would allow those 
fluent in it to use the grammar to flexibly describe all
sorts of graphics, which a computer could then **render**
into the desired visual representation.  Indeed,
the core ideas contained in Wilkerson's grammar of 
graphics have been used to design and implement
a number of computer graphics systems including
... 

In some of these systems, the code resembles the grammar
especially closely.  Mosaic, and hence {vgplotr}, is
one of those systems.

Important components of the grammar of graphics (using
terminology to match {vgplotr}) include

marks (geoms, glyphs)
: The visual elements (dots, lines, etc.) that appear
on the graphic

attributes (properties, aesthetics)
: Each mark has a number of attributes 
(e.g., position, color, size, etc.)
that determine how and where it is displayed within
the graphic.

scales
: A scale is a function that **maps** data values to 
visual values of attributes 
(position, color, shape, etc.) that 
determine how and where those data values are represented
on the graphic.  Attribute values can also be 
**set** to constant values that do not depend on data.

guides
: A guide is a visual element that helps the reader of
the graphic reverse the scale, i.e., determine a 
data value corresponding to the value of a visual
attribute.

transformations
: A transformation is a function applied to data
before visualization.  A histogram, for example,
uses a transformation that puts data values into 
bins and counts how many values are in each bin.

A **specification** describes the marks, attributes,
scales, guides, and transformations and how they are 
to be combined to produce a graphic.

Many graphics consist of several (sub)**plots** that
are arranged or layered to produce the composite 
graphic.

# Code conventions used in {vgplotr}

## snake_case

Various mosaic APIs use either camelCase or snake_case
for names of elements of a mosaic specification.
{vgplotr} adopts snake_case (as does the Python API).
To translate camelCase to snake_case, simply insert
an underscore before any block of capitals and then
make all letters lower case as in the examples below.

```{r}
#| label: camel-to-snake
#| echo: false

tribble(
  ~camelCase    , ~snake_case    ,
  "fillOpacity" , "fill_opacity" ,
  "scaleX"      , "scale_x"
) |>
  gt()
```

This makes it easy to translate documentation written
for one naming style into the other.

Nearly all functions in {vgplotr} begin `vg_`.
Functions for specifying marks begin `vg_mark_`,
for scales, `vg_scale_`, for guides `vg_guide_`, etc.
Generally, the names resemble the names used 
in mosaic, but there are a few important exceptions.

* mosaic uses axis or legend for guides, depending on 
which type of guide it is. This leads to some awkward
name collisions for axes which are easily avoided 
by using *guide* for both axes and legends.

## Formulas for mapping

Attributes can be mapped to data or set to constant values.
To map an attribute to a variable, we use a formula. This 
reduces the need for quoting or non-standard evaluation.

## Use of the pipe (`|>`)

The first argument for most `vg_*()` functions
is `spec` (a specifcation for a graphic or a graphic 
fragment), and most also return a specification.
This allows a piped workflow to build up the components
of a graphic specification.

To make simple graphics easier to specify, `vg_mark_*()`
functions also allow `spec` to be a data source (e.g., a 
data frame).  This allows us to pipe a data frame -- perhaps
after some tidyverse data wrangling -- directly into
the first mark of a layered graphic.

```{r}
#| label: spec-as-data
library(vgplotr)
library(palmerpenguins)

penguins |>
  filter(!is.na(sex)) |>
  filter(species == "Gentoo") |>
  vg_mark_dot(
    x = ~bill_depth_mm,
    y = ~bill_length_mm,
    fill = ~sex
  ) |>
  vg_mark_regression_y(
    x = ~bill_depth_mm,
    y = ~bill_length_mm,
    stroke = ~sex
  ) |>
  vg_guide_color(label = "sex") |>
  vg_meta(title = "Gentoo penguins") |>
  # vg_render() |>
  to_yaml(suppress_data = TRUE) |>
  cat()
```

