---
title: "vgplotr Design Principles"
vignette: >
  %\VignetteIndexEntry{vgplotr Design Principles}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


```{r}
#| label: setup
#| include: false
library(tibble)
library(dplyr)
library(gt)
library(vgplotr)
library(palmerpenguins)
```

Part of learning a new graphics system is learning the 
principles that were used in its design.  This
vignette lays out some of the important vocabulary,
concepts, and design principles used to create vgplotr,
many of which are inherited either from mosaic or
from the grammar of graphics more generally.

# What is a graphic?  What is a grammar of graphics?

A **graphic** is a visual^[Perhaps *sensory* would be 
better than visual, since once could make tactile
representations of data as well. But since vgplotr
is concerned with *visual* representations, we will
stick with visual here.] 
display of data.  Leland Wilkerson proposed a 
**grammar of graphics** that would allow those 
fluent in it to use the grammar to flexibly describe all
sorts of graphics, which a computer could then **render**
into the desired visual representation.  Indeed,
the core ideas contained in Wilkerson's grammar of 
graphics have been used to design and implement
a number of computer graphics systems including
the iNVizN system Wilkinson developed for SPSS,
Tableu (where Wilkinson lead the graphics team),
ggplot2 for graphics in R, and 
vega-lite, which is familiar to Python users via the altair package.
In some of these systems, the code resembles the grammar
especially closely.  Mosaic, and hence vgplotr, is
one of those systems.

Instead of focusing on providing 
a fixed list of graphic types (histograms, scatter plots, etc.),
the grammar of graphics approach focus on providing 
structures, operations, and operations that can be flexibly
combined to form many types of graphics.

For those familiar with a graphics tool based on the grammar of 
graphics, these concepts will be familiar, although the terminology
used in Mosaic may differ.  As one example of terminology 
differences, what the original grammar of graphics
calls geometric objects and {ggplot2} calls geoms, Mosaic calls
marks.

Below is a list of 
important components of the grammar of graphics,
using terminology to match vgplotr (with other commonly used
terminology in parentheses).

graphic
: The entire collection of elements that make up a data visualization.

plot (subplot, panel)
: A rectangular (in Cartesian coordinates) portion of a graphic.
Plots (and other components like interactive controls and data tables) 
can be concatenated horizontally or vertically to create complex
graphics.

mark (geom, glyph)
: The visual element (dots, lines, etc.) that appear
on the graphic. Using multiple marks in the same plot
produces a layered plot.

attribute (property, aesthetic)
: Each mark has a number of attributes 
(e.g., position, color, size, etc.)
that determine how and where it is displayed within
the graphic.

scale
: A scale is a function that **maps** data values to 
visual values of attributes 
(position, color, shape, etc.) that 
determine how and where those data values are represented
on the graphic.  Attribute values can also be 
**set** to constant values that do not depend on data.

guide (axis or legend)
: A guide is a visual element that helps the reader of
the graphic reverse the scale, i.e., determine a 
data value corresponding to the value of a visual
attribute.

transformation (statistics)
: A transformation is a function applied to data
before visualization.  A histogram, for example,
uses a transformation that puts data values into 
bins and counts how many values are in each bin.

interactor
: An interactor is a component of a graphic that viewers
can interact with to change some aspect of the graphic.
Examples: menu, search bar, slides.

A **specification** describes the marks, attributes,
scales, guides, transformations, etc. and how they are 
to be combined to produce a graphic.  Mosaic specifications
can be written in JSON, YAML, javascript, or Python.
vgplotr proivdes a way to produce specifications in R.


# Code conventions used in vgplotr

## snake_case

Various mosaic APIs use either camelCase or snake_case
for names of elements of a mosaic specification.
vgplotr adopts snake_case (as does the Python API).
To translate camelCase to snake_case, simply insert
an underscore before any block of capitals and then
make all letters lower case as in the examples below.

```{r}
#| label: camel-to-snake
#| echo: false

tribble(
  ~camelCase    , ~snake_case    ,
  "fillOpacity" , "fill_opacity" ,
  "scaleX"      , "scale_x"
) |>
  gt()
```

This makes it easy to translate documentation written
for one naming style into the other.

Nearly all functions in vgplotr begin `vg_`.
Functions for specifying marks begin `vg_mark_`,
for scales, `vg_scale_`, for guides `vg_guide_`, etc.
Generally, the names resemble the names used 
in Mosaic, but there are a few important exceptions.

* Mosaic uses axis or legend for guides, depending on 
which type of guide it is. This leads to some awkward
name collisions for axes (which can also be a mark)
which are easily avoided by using *guide* for both axes 
and legends.

## Formulas for mapping

Attributes can be mapped to data or set to constant values.
To map an attribute to a variable, we use a formula. This 
reduces the need for quoting or non-standard evaluation.

## Use of the pipe (`|>`)

The first argument for most `vg_*()` functions
is `spec` (a specification for a graphic or a plot fragment 
fragment), and most also return a specification.
This allows a piped workflow to build up the components
of a graphic specification.

To make simple graphics easier to specify, `vg_mark_*()`
functions also allow `spec` to be a data source (e.g., a 
data frame).  This allows us to pipe a data frame -- perhaps
after some tidyverse data wrangling -- directly into
the first mark of a layered graphic.

```{r}
#| label: spec-as-data
library(vgplotr)
library(palmerpenguins)

penguins |>
  filter(!is.na(sex)) |>
  filter(species == "Gentoo") |>
  vg_mark_dot(
    x = ~bill_depth_mm,
    y = ~bill_length_mm,
    fill = ~sex
  ) |>
  vg_mark_regression_y(
    x = ~bill_depth_mm,
    y = ~bill_length_mm,
    stroke = ~sex
  ) |>
  vg_legend_color(label = "sex") |>
  vg_meta(title = "Gentoo penguins") |>
  # vg_render() |>
  to_yaml(suppress_data = TRUE) |>
  cat()
```

