---
title: "Getting Started with vgplotr"
vignette: >
  %\VignetteIndexEntry{Getting Started with vgplotr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
engine: knitr
---

```{r setup}
library(vgplotr)
```

vgplotr is still very early: there is no JSON/YAML export and no rendering
to an actual mosaic plot yet. What *does* exist is the R-side object model
-- `vg_create()`, `vg_plot()`, marks, interactors, and `param()` -- so this
vignette just builds a few specs and prints their internal structure, as a
sanity check that the pieces fit together the way they're supposed to.

## A single mark

This reproduces the "simple one-layer plot" example from
`design/api-brainstorming.qmd`: one data source, one mark, and a couple of
plot-wide defaults.

```{r}
spec <- vg_create() |>
  vg_data(
    name = "aapl",
    file = "data/stocks.parquet",
    where = "Symbol = 'AAPL'"
  ) |>
  vg_line_y(data_from = "aapl", x = ~Date, y = ~Close) |>
  vg_attributes(width = 680, height = 200)

spec
```

`vg_attributes()` sets top-level `plotDefaults` (`spec$plot_defaults`), not
this plot's own attributes:

```{r}
spec$plot_defaults
spec$layout$attrs
```

## Combining marks and an interactor on one plot

Piping one mark straight into another appends it to the *same* plot,
without needing to wrap them in `vg_plot()` explicitly. Interactors
(`vg_interval_x()` here) can be piped in the same way. This mirrors the
`overview-detail.yaml` example from mosaic's own docs.

```{r}
plot1 <- vg_area_y(data_from = "walk", x = ~t, y = ~v, fill = "steelblue") |>
  vg_interval_x(as = param(brush)) |>
  vg_plot(width = 680, height = 200, name = "focus")

plot1
```

`param(brush)` is how a `$brush`-style reference to a Param/Selection gets
written in valid R -- `$brush` alone isn't parseable:

```{r}
param(brush)
```

## What happens when two marks disagree

Plot-level attributes (like `width`) accumulate silently across marks when
they agree, but a genuine conflict is not resolved silently -- the later
value wins, and you get a warning so it doesn't happen by surprise.

```{r}
plot2 <- vg_dot(x = ~a, y = ~b, width = 400) |>
  vg_line_y(x = ~a, y = ~c, width = 680)

plot2$attrs$width
```

## Not yet implemented

A couple of things you might reach for that aren't built yet:

```{r}
#| error: true
vg_interactor(NULL, "slider", as = param(scale))
```

```{r}
#| error: true
vg_dot(x = ~a, y = ~b) |> vg_vconcat(vg_dot(x = ~a, y = ~c))
```
