---
title: "Getting Started with vgplotr"
vignette: >
  %\VignetteIndexEntry{Getting Started with vgplotr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
engine: knitr
execute: 
  error: true
---

```{r}
#| label: setup
#| include: false
library(vgplotr)
set.seed(12345)
```

## A grammar of interactive graphics

vgplotr provides an R interface to UW Interactive Data Lab's Mosaic/vgplot,
an implementation of a grammar of interactive graphics. To use interactive
features, a graphic must be embedded in an environment that supports
interactive widgets, like an HTML file or a Shiny app.
Non-interactive graphics, supported in other file formats (PDF, Word, etc.),
are created by taking a snapshot of an underlying interactive graphic.

### Installation

vgplotr is designed so that beginners can get up and running quickly by
simply installing the R package and using it in a quarto document.
For now, the package is only available via GitHub:

```{r}
#| eval: false
pak::pak("rpruim/vgplotr")
```

```{r}
#| label: load-vgplotr
library(vgplotr)
```


No additional JavaScript libraries need to be installed by the user.
The Mosaic JavaScript runtime ships with the package.
Mosaic graphics rely on a DuckDB database to store and query the 
underlying data. By default, vgplotr uses DuckDB-Wasm, which 
runs DuckDB in a browser.
DuckDB-Wasm's database engine is an ~35 MB compiled WebAssembly binary
-- too large to include in the package -- so by default
`vg_render()` fetches DuckDB-Wasm
from a CDN each time a graphic is *viewed*.
To avoid this, you can call `vg_cache_duckdb()` once 
to create a local, version-pinned cache and remove the CDN dependency.
Once DuckDB-Wasm has been cached, vgplotr graphics can be rendered
even when you are offline.

```{r}
#| label: cache-duckdb
#| eval: false
vg_cache_duckdb()
```

`vg_info()` provides information about the installation,
including whether DuckDB-Wasm has been cached locally.

```{r}
#| label: duckdb-info
vg_info()
```

:::{.callout-tip}
#### Using native DuckDB

To learn about using native DuckDB (locally or hosted on a server),
see [*Using Databases with vgplotr*](using-databases.qmd).
:::

## Non-interactive graphics

Before getting to interactive elements of graphics, let's begin
by learning how to make graphics without interactive elements.
vgplotr works by creating and then rendering a graphics specification.
A piped workflow can 

* register data sources, 
* add marks (dots, lines, bars, etc.), 
* add interactors (menus, sliders, interval selectors, etc.),
* arrange the layout of the graphic, and 
* set various plot- and graphic-level attributes.

### A graphic with one mark

For our first example, we

* create a graphic specification with a custom title,
* register one data source (a local data frame),
* add one mark (a line mark), and 
* specify the size of our graphic.

```{r}
#| label: aapl-closes
 
stocks <- read.csv("data/stocks.csv")
aapl <- stocks |> dplyr::filter(Symbol == "AAPL")

spec <-
  vg_create(title = "AAPL closing price") |>
  vg_data(name = "aapl", aapl) |>
  vg_mark_line_y(y = ~Close, x = ~Date, data_from = "aapl") |>
  vg_attributes(width = 680, height = 200)

spec
```

`vg_render()` inspects the enclosing environment to determine how the 
graphic can be rendered.

```{r}
#| label: aapl-render
vg_render(spec)
```


### YAML and JSON

vgplotr specs can also be converted to native Mosaic YAML or JSON, 
which can be useful for debugging or for comparing to 
Mosaic documentation. 

:::{.callout-tip}
#### Hiding the data

Since local data is included in the 
specification, it is often handy to hide this with
`suppress_data = TRUE`, especially if the dataset is large.
:::

```{r}
#| label: yaml-spec

spec |> to_yaml(suppress_data = TRUE)
```

<!-- `vg_attributes()` sets attributes directly on the spec's own top level --
distinct from this plot's own attributes (`spec$layout$attrs`, accumulated
from marks/interactors, e.g., via `width =` on `vg_mark_dot()`) and from
mosaic-spec's `plotDefaults` (set via `vg_plot_defaults()`, applied to
*every* plot in the spec -- see the two-plot example below): -->


<!-- spec$attrs
spec$layout$attrs -->

<!-- `vg_create()`'s `title` went to `spec$meta` instead, since it isn't one of
mosaic's plot attributes -- mosaic's own JS runtime never actually reads
`meta` (it's inert, spec-level metadata), but `vg_render()` renders
`meta$title` as a caption above the graphic as a vgplotr-level convenience: -->


### Some shortcuts

We can obtain the same simple plot with less code by taking advantage
of some shortcuts.

* If we pipe a data frame into the `vg_` workflow, we can omit
`vg_create()` and `vg_data()`: `aapl |> vg_mark_line()` is equivalent to
`vg_create() |> vg_data(name = "data1", aapl) |> vg_mark_line()`.

* By default, mark functions use the first registered data source,
  so we can omit `data_from = "data1"`.

* We can set the height and width in `vg_mark_line_y()`.

```{r}
#| label: short-cuts
aapl |>
  vg_mark_line_y(y = ~Close, x = ~Date, width = 680, height = 200) |>
  vg_meta(title = "AAPL closing price") |>
  vg_render()
```

### Multiple marks

Later marks in the same chain don't need to repeat `data_from =` 
if all marks are using the same data source:
any mark that doesn't set its own `data_from`
(and whose mark type requires data) defaults to the spec's
*first* registered data source. 
That makes simple multi-layer plots from
one data source easy to build straight out of a pipe chain:

```{r}
#| label: two-marks
aapl |>
  vg_mark_line_y(y = ~Close, x = ~Date, width = 680, height = 200) |>
  vg_mark_dot(y = ~Close, x = ~Date, fill = "steelblue") |>
  vg_meta(title = "AAPL closing price") |>
  vg_render()
```

To use different data sources for different marks,
name each explicitly with `data_from =`.  The value can be
either a name (e.g., "aapl") or a non-zero integer (e.g., `1L`^[Note: the L here is required -- indices must be explicit integers.])
specifying the registration index.
Negative numbers count from the most recently registered data source.


### Formula shorthand for position channels

Every mark also accepts a `formula` argument -- a shorthand for its position
channels (`x`, `y`, `fx`, `fy`, and paired `x1`/`x2` or `y1`/`y2`) in one
formula instead of naming each channel separately. 
The general format is `y ~ x | fy ~ fx`.
The optional `| fy ~ fx` sets y- and x-faceting.
For marks that require two x-values or two y-values, 
replace `x` or `y` with `x1 + x2` or `y1 + y2`.
Unused channels can be omitted, but since R formulas always require a 
right hand side, use, for example, `a ~ .` to specify `y` or `fy` 
without specifying `x` or `fx`.


```{r}
#| label: formula
mtcars |>
  vg_mark_dot(mpg ~ hp | cyl ~ ., fill = "steelblue", width = 500)
```

is equivalent to writing out every channel by hand:

```{r}
#| label: formula-expanded
mtcars |>
  vg_mark_dot(y = ~mpg, x = ~hp, fy = ~cyl, fill = "steelblue", width = 500)
```

An explicit `x =`/`y =`/etc. alongside `formula`
overrides the value the formula implies for that channel, 
with a warning if the two definitions disagree. 
See `?vg_mark` for the full grammar.

## Loading data from a file

`vg_data()` can be used to register data in a file instead of an
inline data frame. The file can be a file on the file system or it
can be specified as a URL.
Relative paths resolve against wherever the rendered document itself
ends up, the same way a relative link or image would in an ordinary
web page.

:::{.callout-tip}
#### Accessing Mosaic example data sets

`vg_example_url()` returns the URL for a Mosaic example dataset
from its name.

:::

```{r}
#| label: vg-data-url
vg_create() |>
  vg_data(name = "weather", file = vg_example_url("seattle-weather.parquet")) |>
  vg_mark_dot(temp_max ~ date, fill = "steelblue") |>
  vg_render()
```

:::{.callout-warning}
#### One DuckDB instance for an entire *page*.
Every `vg_render()` call on the same page
shares one DuckDB instance behind the scenes, so data sources only
need to be registered once.  This also means that it is not
possible to use the same name for different data sources
in different specifications on the same page. This includes the
implicit `"data1"`/`"data2"`/... names used when a data frame is piped
directly into a mark (see "Some shortcuts" above).

Good practice is to explicity name data sources you plan to 
reuse with distinct, meaningful names (not `data1`, `data2`, etc.).
If a name gets reused, only the last use will remain in the 
database, and any plots that rely on a previous data source
assigned to that name will be incorrect or fail to render altogether
(if the data source doesn't include information the graphic needs).

`vg_data_registry()` lists every name
registered so far this session and what it currently refers to -- a
quick way to check for this before rendering, since nothing else will
point out the mistake.
:::

### Using a subset of a data source

We can obtain a subset of the data contained in the file using
`where =` to declare the WHERE clause of an SQL query.


```{r}
#| label: data-in-file
vg_create() |>
  vg_data(
    name = "aapl_file",
    file = "data/stocks.csv",
    where = "Symbol = 'AAPL'"
  ) |>
  vg_mark_line_y(data_from = "aapl_file", x = ~Date, y = ~Close) |>
  vg_attributes(width = 680, height = 200) |>
  vg_render()
```




## Graphics with multiple plots

In the lingo of Mosaic/vgplot, a plot is a rectangular region 
of a graphic that may contain one or more marks. In other graphics
systems, these might be called panels or sub-plots. 
We can use `vg_hconcat()` and `vg_vconcat()` to place plots 
(and interactive elements) next to each other.

```{r}
#| label: concat

lines <- 
  vg_mark_line_y(
    Close ~ Date, data_from = "aapl", 
    width = 600, height = 150)
dots <- 
  vg_mark_dot(
    Close ~ Date, data_from = "aapl", 
    width = 600, height = 150)

vg_create() |>
  vg_data(name = "aapl", aapl) |>
  vg_attributes(width = 600, height = 300) |>
  vg_vconcat(lines, dots) |>
  vg_render()
```

## Interaction

### Interactors

Piping one mark straight into another appends it to the *same* plot,
without needing to wrap them in `vg_plot()` explicitly. Interactors
(`vg_interval_x()` here) can be piped in the same way.
If you brush over the graphic below, a rectangle will appear to show
the area that has been brushed.  The selection interval of x-values from 
the brush is stored in a parameter (`param(brush)`).
For the moment, nothing responds to this parameter;
we'll add that in a moment.

```{r}
#| label: interval

overview <-
  vg_mark_area_y(
    formula = temp_max ~ date,
    data_from = "weather",
    fill = "steelblue",
    width = 600,
    height = 100
  ) |>
  vg_interval_x(as = param(brush))

vg_create() |>
  vg_data(name = "weather", file = vg_example_url("seattle-weather.parquet")) |>
  vg_plot(overview) |>
  vg_render()
```

<!-- `param(brush)` is how a `$brush`-style reference to a Param/Selection gets
written in valid R -- `$brush` alone isn't parseable: -->

```{r}
#| label: param-brush
#| include: false
param(brush)
```


### Two linked plots with vg_vconcat()

Now we create another version of our plot that responds to
the brush by filtering the data (and zooming in on that data).


```{r}
detail <-
  vg_mark_area_y(
    temp_max ~ date,
    data_from = "weather",
    fill = "steelblue",
    filter_by = param(brush),
    width = 600,
    height = 300
  )

vg_create() |>
  vg_data(name = "weather", file = vg_example_url("seattle-weather.parquet")) |>
  vg_attributes(width = 600, height = 400) |>
  vg_vconcat(detail, overview) |>
  vg_render()
```

Some additional example of interaction appear below.

## Transform functions: binning and aggregation

Mosaic does not have a histogram "mark" (or an other histogram primative).
From the perspective of Mosaic, a histogram is a graphic with bar marks
created from binned counts.  So we need to specify three things to 
make a histogram:

* the bar mark,
* what to bin, and 
* that we want to count how many items are in each bin.

Channel encoding formulas can use transform functions like 
`vg_bin()`/`vg_count()`
(mirroring mosaic's own `bin`/`count` transforms) instead of a bare column
reference. The binning and counting for the histogram below are all done 
in DuckDB, not in R.  

```{r}
delays <-
  data.frame(delay = rgamma(2000, shape = 2, rate = 0.3) - 5)

vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_mark_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_attributes(width = 680, height = 200) |>
  vg_render()
```

Transform functions call also be called directly, 
useful for inspecting what it builds:

```{r}
vg_bin(delay, step = 2)
```

## More marks

vgplotr includes a `vg_mark_*()` function for each of the 65
marks supported by Mosaic.
Each of these functions recognizes every plot-level attribute 
(215) as belonging to the plot, not the mark -- 
even when a mark has a same-named property of its own 
(e.g., a mark's own `inset` vs. the plot-wide default
`inset`, told apart by `vg_mark_*()`'s generated list of each 
mark's real properties).


```{r}
vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_mark_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_mark_rule_x(x = 0, stroke = "firebrick", stroke_width = 2) |>
  vg_plot(
    width = 680,
    height = 200,
    margin_left = 60,
    aspect_ratio = 1
  ) |>
  vg_render()
```

## Scales and guides

**Scales** determine how data values are converted into plot values.
For example, if we use color to indicate species, it is the job of a 
scale to determine which color is used for which species, and we modify
the scale to get a different color scheme.

**Guides** (axes and legends) indicate to the viewer of the graphic
how to reverse the scale and convert plot values back into data values.
So a guide (legend) for species color, shows how to convert a displayed 
color back into a species.

The graphic below modifies the default scales and guides to provide

* a logarithmic scale for y;
* custom labels for the x- and y-axes;
* grid lines for x.


```{r}
vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_mark_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_scale_y(type = "symlog") |>
  vg_guide_x(label = "Delay (minutes)", grid = TRUE) |>
  vg_guide_y(label = "Count") |>
  vg_plot(width = 680, height = 200) |>
  vg_render()
```

`vg_scale_fx()`/`vg_scale_fy()` and `vg_guide_fx()`/`vg_guide_fy()` do the
same for facets. 
Facet scales are always band scales, so `vg_scale_fx()`/`vg_scale_fy()`
expose a smaller set of properties than `vg_scale_x()`/`vg_scale_y()` (no
`type`, `nice`, `zero`, or `clamp`).


```{r}
vg_create() |>
  vg_data(name = "stocks", file = vg_example_url("stocks.csv")) |>
  vg_mark_line(Close ~ Date | Symbol ~ ., data_from = "stocks") |>
  vg_scale_fy(padding = 0.1) |>
  vg_guide_fy(label = "Ticker Symbol") |>
  vg_mark_frame() |>
  vg_plot(width = 680, height = 400) |>
  vg_render()
```

:::{.callout-tip}
#### vg_mark_frame()

By default, Mosaic plots do not include a boundary rectangle. For faceted
graphics, such a frame can often make it easier to see the facets.
`vg_mark_frame()` can be added to non-faceted plots as well.
:::

:::{.callout-warning}
#### Missing facet features

Those familiar with `ggplot2` or Vega-Lite may wonder about

* wrapped facets,
* independent scales in facets.

Neither of these is currently available in Mosaic (and hence not in
vgplotr). Independent scales can be emulated by manually filtering
and concatenating.
Facet wrapping can be emulated by manually creating 
row and column variables and using those to facet.

:::


### Other scales and their guides

`vg_scale_color()`/`vg_scale_opacity()`/`vg_scale_r()` (aliased
`vg_scale_radius()`) and `vg_guide_color()`/`vg_guide_opacity()`/
`vg_guide_r()` (aliased `vg_guide_radius()`) do the same for the `color`,
`opacity`, and `r` (radius) channels -- the scales that `fill`/`stroke`,
`opacity`, and `r` encodings are bound to. 

### Domain and range

The domain and range of a scale specify the inputs and outputs of the scale.
So 
```r
  vg_scale_color(
    domain = c("A", "B", "C", "D"), 
    range = c("red", "darkgreen", "steelblue", "orange")
```

below instructs the color scale to show group A as red, group B as green, etc.

```{r}
pts <- data.frame(
  x = rnorm(90),
  y = rnorm(90),
  group = sample(LETTERS[1:4], 90, replace = TRUE),
  size = runif(90, 1, 10)
)

pts |>
  vg_mark_dot(y ~ x, fill = ~group, r = ~size, fill_opacity = 0.6, data_from = "pts") |>
  vg_scale_color(
    domain = c("A", "B", "C", "D"),
    range = c("red", "darkgreen", "steelblue", "orange")
  ) |>
  vg_scale_r(range = c(2, 12)) |>
  vg_guide_color(label = "Group") |>
  vg_guide_r(label = "Size") |>
  vg_plot(width = 400, height = 300) |>
  vg_render()
```

<!-- These are deliberately named `vg_guide_*()` rather than `vg_legend_*()`:
`vg_legend_color()`/`vg_legend_opacity()` already exist (next section) and
mean something different -- adding an actual rendered legend mark to the
spec. `vg_guide_color()`/`vg_guide_opacity()`/`vg_guide_r()` only set the
`colorLabel`/`opacityLabel`/`rLabel` (and `colorTickFormat`/
`opacityTickFormat`) plot attributes; they neither add nor require a
legend. -->

### Setting attributes for multiple scales at once

`vg_scale_all()`/`vg_guide_all()` set mosaic's plot-wide *default* scale/
guide properties -- `align`/`inset`/`padding` and `axis`/`grid`/
`ariaLabel`/`ariaDescription` respectively -- which apply to every
position scale/axis (`x`, `y`, `fx`, `fy`) that doesn't set its own value.
Unlike `vg_scale_x(grid = TRUE)`/`vg_guide_y(grid = TRUE)`/..., which only
affect one axis, `vg_guide_all(grid = TRUE)` turns grid lines on for both
axes at once, without repeating the call per axis (a later
`vg_guide_x(grid = FALSE)` would still override it just for `x`, since the
per-axis attribute always wins over the default).

`vg_scale_position()` and `vg_scale_facet()` are similar, but affect only
the x- and y- or fx- and fy-scales, respectively.


```{r}
pts |>
  vg_mark_dot(y ~ x, fill = ~group, fill_opacity = 0.6, data_from = "pts") |>
  vg_scale_all(padding = 0.1) |>
  vg_guide_all(grid = TRUE) |>
  vg_plot(width = 400, height = 300) |>
  vg_render()
```

:::{.callout-warning}
#### Axis marks

In addition to axis scales and guides, Mosaic provides axis *marks*.
These are available as 
`vg_mark_axis_x()`/ `vg_mark_axis_y()` and allow the user to 
create separately drawn, independently-styled axes.
:::


### Legends

By default, axes appear for the x- and y-axes and for facets.
Legends, on the other hand, must be inserted manually.
A legend embedded in a plot (piped in alongside its marks, with no
`for_plot`) picks up a matching encoding automatically -- 
here, the color mapping from `fill = ~group`.

```{r}
pts |>
  vg_mark_dot(y ~ x, fill = ~group, r = ~size, fill_opacity = 0.6, data_from = "pts") |>
  vg_legend_color() |>
  vg_plot(width = 400, height = 300) |>
  vg_render()
```

Alternatively, we can place legends with `vg_hconcat()` or `vg_vconcat()`.

```{r}
vg_create() |>
  vg_data(data = pts, name = "points") |>
  vg_vconcat(
    vg_mark_dot(
      name = "dots",
      data_from = "points",
      y ~ x,
      fill = ~group,
      r = ~size,
      fill_opacity = 0.6
    ),
    vg_legend_color(for_plot = "dots")
  ) |>
  vg_render()
```

For guides that appear as legends, `vg_guide_*()` controls *how* the
legend appears and `vg_legend_*()` controls *where* (and *whether*) 
the legend appears.


## Layout-level inputs

Inputs like `vg_menu()`/`vg_slider()` are standalone widgets that live
alongside plots in the layout (via `vg_vconcat()`/`vg_hconcat()`), rather
than inside one. Here a menu drives a `param()` that's used directly as a
mark's `fill` (no formula needed, since it isn't a column mapping) -- pick a
different color and the plot updates.

```{r}
color_picker <- vg_menu(
  label = "Color",
  as = param(dotcolor),
  options = c("steelblue", "orange", "firebrick"),
  value = "firebrick"
)

colored_plot <-
  vg_mark_dot(data_from = "pts", y ~ x, fill = param(dotcolor)) |>
  vg_plot(width = 400, height = 300)

vg_create() |>
  vg_data(name = "pts", data = pts) |>
  vg_vconcat(color_picker, colored_plot) |>
  vg_render()
```

## What happens when two marks disagree

Plot-level attributes (like `width`) accumulate silently across marks when
they agree, but a genuine conflict is not resolved silently -- the later
value wins, and you get a warning so it doesn't happen by surprise.

```{r}
conflicted_graphic <-
  vg_mark_dot(x = ~a, y = ~b, width = 400) |>
  vg_mark_line_y(x = ~a, y = ~c, width = 680)

conflicted_graphic$attrs$width
```

## `sql()`/`agg()`: raw SQL expressions

`sql()` (and `agg()`, for expressions with an aggregate function) wraps a
raw SQL expression for encodings that transforms and plain column
references can't express -- mosaic-spec's own `{sql: "..."}`/`{agg: "..."}`.
Below, a menu-controlled parameter chooses a temperature scale, and 
an SQL expression is computed by DuckDB to convert the temperature from 
Celsius to Fahrenheit as needed.

Two things to note here:

*   In the SQL expression parameter names are preceded with `$`. 
    (`param(temp_unit)` does almost the same thing but escaping appropriately
    since R variable names cannot begin with `$` unless escaped.)

    ```{r}
    param(temp_unit)
    ```

*   SQL expressions are not allowed everywhere in the Mosaic specification.
In particular, they are not allowed in the `label =` of `vg_guide_y()`.
We have worked around this by having the value of the parameter be the
text we want, since a parameter value is allowed in this context.

```{r}
#| label: sql
#| 
temp_unit_picker <- vg_menu(
  label = "Temperature scale",
  as = param(temp_unit),
  options = c("Temperature (°C)", "Temperature (°F)"),
  value = "Temperature (°C)"
)

bilingual_plot <-
  vg_mark_dot(
    data_from = "weather",
    x = ~date,
    y = sql(
      "CASE WHEN $temp_unit = 'Temperature (°C)' THEN temp_max ELSE (temp_max * 9.0/5.0 + 32) END"
    ),
    fill = "steelblue"
  ) |>
  vg_guide_y(label = param(temp_unit)) |>
  vg_guide_x(label = NA) |>
  vg_plot(width = 600, height = 200)

vg_create() |>
  vg_data(name = "weather", file = vg_example_url("seattle-weather.parquet")) |>
  vg_vconcat(bilingual_plot, vg_vspace("1em"), temp_unit_picker) |>
  vg_render(height = 250, width = 600)
```

:::{.callout-tip}
#### Adding space to the layout

We can use `vg_hspace()` and `vg_vspace()` to add additional space between
concatenated elements.  Without the extra space in the example above,
the menu is uncomfortably close to the plot.
:::

## `to_json()`/`to_yaml()`: converting a spec to mosaic's own format

`vg_render(spec)` renders a spec directly, but sometimes what's wanted is
the underlying mosaic-spec JSON or YAML itself -- to inspect exactly what
an R-side spec translates to, or to save/share it independent of R (e.g.,
to feed to mosaic's own JS `parseSpec()`, or as a `.json`/`.yaml` file
alongside other mosaic examples). `to_json()`/`to_yaml()` work on a full
`vgspec` or, as here, on a bare fragment that was never wrapped in
`vg_create()` -- note that `param(brush)` becomes mosaic's own `"$brush"`
string:

```{r}
frag <-
  vg_mark_dot(x = ~x, y = ~y, fill = "steelblue") |>
  vg_interval_x(as = param(brush))

to_json(frag)
```

```{r}
to_yaml(frag)
```

A data frame passed to `vg_data()` becomes an inline array of row
objects (mosaic-spec's own `{data: [...]}` form), matching how mosaic's
own example specs represent small inline datasets:

```{r}
#| label: to-yaml
vg_create() |>
  vg_data(name = "aapl", data = aapl) |>
  vg_mark_line_y(data_from = "aapl", x = ~Date, y = ~Close) |>
  to_yaml()
```

:::{.callout-tip}
#### Hiding the data

If you don't want to include the inline definition of the data,
use `suppress_data = TRUE`:

```{r}
#| label: to-yaml-no-data
vg_create() |>
  vg_data(name = "aapl", data = aapl) |>
  vg_mark_line_y(data_from = "aapl", x = ~Date, y = ~Close) |>
  to_yaml(suppress_data = TRUE)
```

:::

## Rendering a spec from elsewhere

`vg_render()` also accepts a JSON or YAML string directly -- an
already-complete mosaic spec, e.g., copied from mosaic's own [example
gallery](https://idl.uw.edu/mosaic/examples/) or the output of
`to_json()`/`to_yaml()` above -- so an example found elsewhere can be
dropped straight into a Quarto document without translating it into
`vg_*()` calls first. Format is auto-detected from the text:

```{r}
vg_render('
meta:
  title: Pasted from elsewhere
data:
  penguins:
    file: https://raw.githubusercontent.com/uwdata/mosaic/refs/heads/main/data/penguins.parquet
plot:
  - mark: dot
    data:
      from: penguins
    x: bill_length
    y: bill_depth
    fill: species
')
```

Since it's a plain string, anything it references has to be resolvable
on its own -- a local data frame already loaded into R isn't available
to it the way `vg_data()` makes one available to a `vgspec`, so
`file =` needs to be a real, fetchable URL (as above) rather than a
local path, unless the rendered page will always be opened from that
same local directory.

Every example in this vignette renders through DuckDB-Wasm -- a real
DuckDB compiled to WebAssembly and run entirely inside the browser, which
is what keeps these graphics self-contained and shareable as plain HTML.
For data too large for the browser, or that actually lives in another
database (Postgres, MySQL, SQLite, via DuckDB's own federation
extensions), see `vignette("using-databases")` for `vg_duckdb_connector()`,
an opt-in way to render against a real, native DuckDB process instead.
