---
title: "Getting Started with vgplotr"
vignette: >
  %\VignetteIndexEntry{Getting Started with vgplotr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
engine: knitr
---

```{r setup}
library(vgplotr)
```

vgplotr is still very early, but it can now render a live mosaic vgplot:
this vignette builds a couple of specs with the R-side object model --
`vg_create()`, `vg_plot()`, marks, interactors, and `param()` -- prints
their internal structure as a sanity check, and then renders them with
`vg_render()`.

**Rendering mostly works offline already.** The mosaic JS runtime itself
ships with the package, so no CDN is needed for that. The one exception is
DuckDB's actual database engine (a ~35 MB compiled WebAssembly binary, too
large to include in the package) -- by default `vg_render()` fetches that
one piece from a CDN each time a plot is viewed, unless you run
`vg_cache_duckdb()` once, which downloads it into a local, version-pinned
cache and removes the CDN dependency entirely (`vg_render()` will offer to
do this for you interactively the first time, if you don't already have it
cached). `vg_render()`'s `use_cache` argument tracks the cache automatically,
but can override it for a single plot -- `use_cache = FALSE` forces that one
plot to use the CDN even if a cache exists.

## A single mark

This uses a small inline data frame passed directly to `vg_data()` --
see the next section for the file-based version.

```{r}
aapl <- data.frame(
  Date = c(
    "2020-01-01",
    "2020-02-01",
    "2020-03-01",
    "2020-04-01",
    "2020-05-01",
    "2020-06-01"
  ),
  Close = c(296.24, 313.05, 254.29, 293.80, 317.94, 364.80)
)

spec <-
  vg_create(title = "AAPL closing price") |>
  vg_data(name = "aapl", data = aapl) |>
  vg_mark_line_y(data_from = "aapl", x = ~Date, y = ~Close) |>
  vg_attributes(width = 680, height = 200)

spec
```

`vg_attributes()` sets attributes directly on the spec's own top level --
distinct from this plot's own attributes (`spec$layout$attrs`, accumulated
from marks/interactors, e.g. via `width =` on `vg_mark_dot()`) and from
mosaic-spec's `plotDefaults` (set via `vg_plot_defaults()`, applied to
*every* plot in the spec -- see the two-plot example below):

```{r}
spec$attrs
spec$layout$attrs
```

`vg_create()`'s `title` went to `spec$meta` instead, since it isn't one of
mosaic's plot attributes -- mosaic's own JS runtime never actually reads
`meta` (it's inert, spec-level metadata), but `vg_render()` renders
`meta$title` as a caption above the plot as a vgplotr-level convenience:

```{r}
spec$meta
```

```{r}
vg_render(spec)
```

### A shortcut: piping a data frame straight into a mark

For a quick plot -- especially at the end of a data-wrangling pipe chain,
where there's no `vg_create() |> vg_data(...)` step to hang the data off
of yet -- a data frame can be piped directly into a mark instead of
`spec`. `some_data |> vg_mark_dot(...)` is shorthand for `vg_create() |>
vg_data(data = some_data) |> vg_mark_dot(data_from = "data", ...)`: it
starts a new spec, registers the data frame (named `"data"` by default --
see [vg_data()] for the naming rule, and pass `data_from =` explicitly to
choose the name yourself), and wires it up as the mark's own `data_from`
automatically.

```{r}
aapl |> vg_mark_line_y(x = ~Date, y = ~Close, width = 680, height = 200)
```

Later marks in the same chain don't need to repeat `data_from =` either:
any mark that doesn't set its own `data_from` (and whose mark type takes
data at all -- a handful, like `vg_mark_frame()`/`vg_mark_sphere()`,
compute their own geometry and never take one) defaults to the spec's
*first* registered data source. That makes simple multi-layer plots from
one data source easy to build straight out of a pipe chain:

```{r}
aapl |>
  vg_mark_line_y(x = ~Date, y = ~Close) |>
  vg_mark_dot(x = ~Date, y = ~Close, fill = "steelblue") |>
  vg_plot(width = 680, height = 200)
```

For more than one data source, name each explicitly with `data_from =`
as usual -- the default only ever reaches for the *first* one registered.
`data_from =` also accepts a nonzero 1-based integer *index* into the
spec's data sources instead of a name, negative counting from the end
(`data_from = -1L` for the most recently registered source) -- note the
`L`: a plain number like `data_from = 0` keeps its existing meaning,
mosaic's own literal inline-data shorthand (`"data": [0]`, e.g. for a
reference line), so only a real R integer is read as an index --
`data_from = 0L` included, since `0` is never a valid index anyway.

## Loading data from a file

`vg_data()` can also point at a file instead of an inline data frame -- the
fictional `data/stocks.parquet` from the original design doc, made real as
a small `.csv` file shipped alongside this vignette. A relative path like
this resolves against wherever the rendered document itself ends up, the
same way a relative link or image would in an ordinary web page.

(Named `"aapl_file"` here, not `"aapl"` -- every `vg_render()` call on this
page shares one DuckDB instance behind the scenes, so two different specs
loading different data under the same name would collide.)

```{r}
spec1b <-
  vg_create() |>
  vg_data(
    name = "aapl_file",
    file = "data/stocks.csv",
    where = "Symbol = 'AAPL'"
  ) |>
  vg_mark_line_y(data_from = "aapl_file", x = ~Date, y = ~Close) |>
  vg_attributes(width = 680, height = 200)

vg_render(spec1b)
```

## Combining marks and an interactor on one plot

Piping one mark straight into another appends it to the *same* plot,
without needing to wrap them in `vg_plot()` explicitly. Interactors
(`vg_interval_x()` here) can be piped in the same way.

```{r}
walk <- data.frame(t = 1:100, v = cumsum(rnorm(100)))

spec2 <-
  vg_create() |>
  vg_data(name = "walk", data = walk) |>
  vg_mark_area_y(data_from = "walk", x = ~t, y = ~v, fill = "steelblue") |>
  vg_interval_x(as = param(brush)) |>
  vg_plot(width = 680, height = 200, name = "focus")

spec2
```

`param(brush)` is how a `$brush`-style reference to a Param/Selection gets
written in valid R -- `$brush` alone isn't parseable:

```{r}
param(brush)
```

```{r}
vg_render(spec2)
```

## Two linked plots with vg_vconcat()

This reproduces mosaic's own `overview-detail.yaml` example: an "overview"
plot with a brush interactor, stacked above a "detail" plot that's filtered
by whatever the brush selects. Drag a selection on the top plot to see the
bottom one update.

Passing separate mark/interactor chains to `vg_vconcat()` stacks them as
*separate* plots -- unlike piping them together, which (as above) keeps them
on one shared plot. `vg_plot_defaults()` sets mosaic-spec's `plotDefaults`,
applied to *every* plot in the spec -- here, both `overview` and `detail`
get `width`/`height` without repeating them on each `vg_plot()` call:

```{r}
overview <-
  vg_mark_area_y(data_from = "walk", x = ~t, y = ~v, fill = "steelblue") |>
  vg_interval_x(as = param(brush)) |>
  vg_plot()

detail <-
  vg_mark_area_y(
    data_from = "walk",
    filter_by = param(brush),
    x = ~t,
    y = ~v,
    fill = "steelblue"
  ) |>
  vg_plot()

spec3 <-
  vg_create() |>
  vg_data(name = "walk", data = walk) |>
  vg_vconcat(overview, detail) |>
  vg_plot_defaults(width = 680, height = 150)

spec3
```

```{r}
vg_render(spec3)
```

## Transform functions: binning and aggregation

Encoding formulas can use transform functions like `vg_bin()`/`vg_count()`
(mirroring mosaic's own `bin`/`count` transforms) instead of a bare column
reference. This reproduces the histogram pattern from mosaic's
`flights-200k.yaml` example (`x: {bin: delay}`, `y: {count:}`), binning a
continuous variable and counting rows per bin -- all computed by DuckDB, not
in R.

```{r}
delays <- data.frame(delay = rgamma(2000, shape = 2, rate = 0.3) - 5)

spec4 <-
  vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_mark_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_attributes(width = 680, height = 200)

spec4
```

Every transform function also works called directly, without a formula --
useful for inspecting what it builds:

```{r}
vg_bin(delay, step = 2)
```

```{r}
vg_render(spec4)
```

## Generated marks and attributes

vgplotr now has a wrapper for every mark type mosaic supports (65, generated
from mosaic's own JSON schema by `data-raw/update-schema.R` -- see
`vg_mark_hexbin()`, `vg_mark_text()`, `vg_mark_rule_x()`, ... alongside the marks used
above), and correctly recognizes every plot-level attribute (215) as
belonging to the plot, not the mark -- even when a mark has a same-named
property of its own (e.g. a mark's own `inset` vs. the plot-wide default
`inset`, told apart by `vg_mark_*()`'s generated list of each mark's real
properties).

Every mark/interactor argument is snake_case (`stroke_width =`, not
`strokeWidth =`, above), and so is every plot-level attribute passed to
`vg_plot()`/`vg_attributes()`/`vg_plot_defaults()`/`vg_create()`
(`margin_left =`, not `marginLeft =`, below) -- all generated straight from
mosaic's schema and forwarded under the exact camelCase key mosaic's JSON
spec needs (`strokeWidth`, `marginLeft`) under the hood, so this is purely
a naming convention on the R side.

```{r}
spec4b <-
  vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_mark_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_mark_rule_x(x = 0, stroke = "firebrick", stroke_width = 2) |>
  vg_plot(width = 680, height = 200, margin_left = 60, aspect_ratio = 1)

vg_render(spec4b)
```

## Scales and axis guides

`vg_scale_x()`/`vg_scale_y()` and `vg_guide_x()`/`vg_guide_y()` are
discoverable, per-channel wrappers around the same plot-level attributes
already set above (`x_domain`, `margin_left`, ...) -- e.g.
`vg_scale_y(type = "sqrt")` is `vg_plot(y_scale = "sqrt")`. Like
`vg_plot()`, they can be piped in anywhere alongside marks and
interactors, since they only ever set attributes on the current plot.

(Note the names: `vg_guide_x()`/`vg_guide_y()`, not `vg_axis_x()`/
`vg_axis_y()` -- those names are used by `vg_mark_axis_x()`/
`vg_mark_axis_y()`, mosaic's `axisX`/`axisY` *mark* (a separately drawn,
independently-styled axis), a different thing from the plain axis-guide
attributes set here.)

```{r}
spec4c <-
  vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_mark_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_scale_y(type = "sqrt") |>
  vg_guide_x(label = "Delay (minutes)", grid = TRUE) |>
  vg_guide_y(label = "Count") |>
  vg_plot(width = 680, height = 200)

vg_render(spec4c)
```

`vg_scale_fx()`/`vg_scale_fy()` and `vg_guide_fx()`/`vg_guide_fy()` do the
same for the facet axes (`fxDomain`, `fxAxis`, ...) -- one plot faceted
into subplots by a grouping column, encoded via `fx =`/`fy =` on a mark.
Facet scales are always band scales, so `vg_scale_fx()`/`vg_scale_fy()`
expose a smaller set of properties than `vg_scale_x()`/`vg_scale_y()` (no
`type`, `nice`, `zero`, or `clamp`).

```{r}
pts2 <- data.frame(
  x = rnorm(120),
  y = rnorm(120),
  group = sample(c("a", "b", "c"), 120, replace = TRUE)
)

spec4d <-
  vg_create() |>
  vg_data(name = "pts2", data = pts2) |>
  vg_mark_dot(data_from = "pts2", x = ~x, y = ~y, fx = ~group) |>
  vg_scale_fx(padding = 0.1) |>
  vg_guide_fx(label = "Group") |>
  vg_plot(width = 680, height = 200)

vg_render(spec4d)
```

`vg_scale_color()`/`vg_scale_opacity()`/`vg_scale_r()` (aliased
`vg_scale_radius()`) and `vg_guide_color()`/`vg_guide_opacity()`/
`vg_guide_r()` (aliased `vg_guide_radius()`) do the same for the `color`,
`opacity`, and `r` (radius) channels -- the scales that `fill`/`stroke`,
`opacity`, and `r` encodings are bound to. These three
channels aren't a symmetric family the way `x`/`y` are, so each gets its
own dedicated argument list rather than a shared `which =` generic --
e.g. `vg_scale_color()` has palette-specific arguments
(`scheme =`, `interpolate =`, `pivot =`, `symmetric =`) that
`vg_scale_opacity()`/`vg_scale_r()` don't, and `vg_scale_r()` has no
`reverse =` (mosaic doesn't define `rReverse`).

```{r}
pts3 <- data.frame(
  x = rnorm(90),
  y = rnorm(90),
  group = rnorm(90),
  size = runif(90, 1, 10)
)

spec4e <-
  vg_create() |>
  vg_data(name = "pts3", data = pts3) |>
  vg_mark_dot(data_from = "pts3", x = ~x, y = ~y, fill = ~group, r = ~size) |>
  vg_scale_color(scheme = "Viridis") |>
  vg_scale_r(range = c(2, 12)) |>
  vg_guide_color(label = "Group") |>
  vg_guide_r(label = "Size") |>
  vg_plot(width = 400, height = 300)

vg_render(spec4e)
```

These are deliberately named `vg_guide_*()` rather than `vg_legend_*()`:
`vg_legend_color()`/`vg_legend_opacity()` already exist (next section) and
mean something different -- adding an actual rendered legend mark to the
spec. `vg_guide_color()`/`vg_guide_opacity()`/`vg_guide_r()` only set the
`colorLabel`/`opacityLabel`/`rLabel` (and `colorTickFormat`/
`opacityTickFormat`) plot attributes; they neither add nor require a
legend.

`vg_scale_all()`/`vg_guide_all()` set mosaic's plot-wide *default* scale/
guide properties -- `align`/`inset`/`padding` and `axis`/`grid`/
`ariaLabel`/`ariaDescription` respectively -- which apply to every
position scale/axis (`x`, `y`, `fx`, `fy`) that doesn't set its own value.
Unlike `vg_scale_x(grid = TRUE)`/`vg_guide_y(grid = TRUE)`/..., which only
affect one axis, `vg_guide_all(grid = TRUE)` turns grid lines on for both
axes at once, without repeating the call per axis (a later
`vg_guide_x(grid = FALSE)` would still override it just for `x`, since the
per-axis attribute always wins over the default).

```{r}
vg_mark_dot(x = ~a, y = ~b) |>
  vg_scale_all(padding = 0.1) |>
  vg_guide_all(grid = TRUE) |>
  vg_plot(width = 400, height = 300)
```

## Legends

A legend embedded in a plot (piped in alongside its marks, with no
`for_plot`) picks up a matching encoding automatically -- here, the color
mapping from `fill = ~group`.

```{r}
pts <- data.frame(
  x = rnorm(90),
  y = rnorm(90),
  group = sample(c("a", "b", "c"), 90, replace = TRUE)
)

spec5 <-
  vg_create() |>
  vg_data(name = "pts", data = pts) |>
  vg_mark_dot(data_from = "pts", x = ~x, y = ~y, fill = ~group) |>
  vg_legend(type = "color") |>
  vg_plot(width = 400, height = 300)

vg_render(spec5)
```

## Layout-level inputs

Inputs like `vg_menu()`/`vg_slider()` are standalone widgets that live
alongside plots in the layout (via `vg_vconcat()`/`vg_hconcat()`), rather
than inside one. Here a menu drives a `param()` that's used directly as a
mark's `fill` (no formula needed, since it isn't a column mapping) -- pick a
different color and the plot updates.

```{r}
color_picker <- vg_menu(
  label = "Color",
  as = param(dotcolor),
  options = c("steelblue", "orange", "firebrick")
)

colored_plot <-
  vg_mark_dot(data_from = "pts", x = ~x, y = ~y, fill = param(dotcolor)) |>
  vg_plot(width = 400, height = 300)

spec6 <-
  vg_create() |>
  vg_data(name = "pts", data = pts) |>
  vg_vconcat(color_picker, colored_plot)

vg_render(spec6)
```

## What happens when two marks disagree

Plot-level attributes (like `width`) accumulate silently across marks when
they agree, but a genuine conflict is not resolved silently -- the later
value wins, and you get a warning so it doesn't happen by surprise.

```{r}
plot3 <-
  vg_mark_dot(x = ~a, y = ~b, width = 400) |>
  vg_mark_line_y(x = ~a, y = ~c, width = 680)

plot3$attrs$width
```

## `sql()`/`agg()`: raw SQL expressions

`sql()` (and `agg()`, for expressions with an aggregate function) wraps a
raw SQL expression for encodings that transforms and plain column
references can't express -- mosaic-spec's own `{sql: "..."}`/`{agg: "..."}`.
Below, a slider-controlled "bias" parameter is added to a plotted value,
computed live by DuckDB. A `param()` reference can be written directly in
the SQL text (`sql("v + $bias")`) or built up from pieces, as below --
both are equivalent.

```{r}
bias_picker <- vg_slider(
  label = "Bias",
  as = param(bias),
  min = -20,
  max = 20,
  step = 1,
  value = 0
)

biased_plot <-
  vg_mark_area_y(
    data_from = "walk",
    x = ~t,
    y = sql("v + ", param(bias)),
    fill = "steelblue"
  ) |>
  vg_plot(width = 680, height = 200, y_domain = c(-30, 30))

spec7 <-
  vg_create() |>
  vg_data(name = "walk", data = walk) |>
  vg_vconcat(bias_picker, biased_plot)

vg_render(spec7)
```

## `to_json()`/`to_yaml()`: converting a spec to mosaic's own format

`vg_render(spec)` renders a spec directly, but sometimes what's wanted is
the underlying mosaic-spec JSON or YAML itself -- to inspect exactly what
an R-side spec translates to, or to save/share it independent of R (e.g.
to feed to mosaic's own JS `parseSpec()`, or as a `.json`/`.yaml` file
alongside other mosaic examples). `to_json()`/`to_yaml()` work on a full
`vgspec` or, as here, on a bare fragment that was never wrapped in
`vg_create()` -- note that `param(brush)` becomes mosaic's own `"$brush"`
string:

```{r}
frag <- vg_mark_dot(x = ~x, y = ~y, fill = "steelblue") |>
  vg_interval_x(as = param(brush))
to_json(frag)
```

```{r}
to_yaml(frag)
```

A data frame passed to `vg_data()` becomes an inline array of row
objects (mosaic-spec's own `{data: [...]}` form), matching how mosaic's
own example specs represent small inline datasets:

```{r}
to_yaml(
  vg_create() |>
    vg_data(name = "aapl", data = aapl) |>
    vg_mark_line_y(data_from = "aapl", x = ~Date, y = ~Close)
)
```

## Rendering a spec from elsewhere

`vg_render()` also accepts a JSON or YAML string directly -- an
already-complete mosaic spec, e.g. copied from mosaic's own [example
gallery](https://idl.uw.edu/mosaic/examples/) or the output of
`to_json()`/`to_yaml()` above -- so an example found elsewhere can be
dropped straight into a Quarto document without translating it into
`vg_*()` calls first. Format is auto-detected from the text:

```{r}
vg_render('
meta:
  title: Pasted from elsewhere
data:
  penguins:
    file: https://raw.githubusercontent.com/uwdata/mosaic/refs/heads/main/data/penguins.parquet
plot:
  - mark: dot
    data:
      from: penguins
    x: bill_length
    y: bill_depth
    fill: species
')
```

Since it's a plain string, anything it references has to be resolvable
on its own -- a local data frame already loaded into R isn't available
to it the way `vg_data()` makes one available to a `vgspec`, so
`file =` needs to be a real, fetchable URL (as above) rather than a
local path, unless the rendered page will always be opened from that
same local directory.
