---
title: "Getting Started with vgplotr"
vignette: >
  %\VignetteIndexEntry{Getting Started with vgplotr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
engine: knitr
---

```{r setup}
library(vgplotr)
```

vgplotr is still very early, but it can now render a live mosaic vgplot:
this vignette builds a couple of specs with the R-side object model --
`vg_create()`, `vg_plot()`, marks, interactors, and `param()` -- prints
their internal structure as a sanity check, and then renders them with
`vg_render()`.

**Rendering mostly works offline already.** The mosaic JS runtime itself
ships with the package, so no CDN is needed for that. The one exception is
DuckDB's actual database engine (a ~35 MB compiled WebAssembly binary, too
large to include in the package) -- by default `vg_render()` fetches that
one piece from a CDN each time a plot is viewed, unless you run
`vg_cache_duckdb()` once, which downloads it into a local, version-pinned
cache and removes the CDN dependency entirely (`vg_render()` will offer to
do this for you interactively the first time, if you don't already have it
cached).

## A single mark

This reproduces the "simple one-layer plot" example from
`design/api-brainstorming.qmd`, using a small inline data frame passed
directly to `vg_data()` -- see the next section for the file-based version
closer to the original example.

```{r}
aapl <- data.frame(
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01", "2020-05-01", "2020-06-01"),
  Close = c(296.24, 313.05, 254.29, 293.80, 317.94, 364.80)
)

spec <-
  vg_create() |>
  vg_data(name = "aapl", data = aapl) |>
  vg_line_y(data_from = "aapl", x = ~Date, y = ~Close) |>
  vg_attributes(width = 680, height = 200)

spec
```

`vg_attributes()` sets top-level `plotDefaults` (`spec$plot_defaults`), not
this plot's own attributes -- with only one plot, `vg_render()` merges the
two, but that will need revisiting once multiple plots exist:

```{r}
spec$plot_defaults
spec$layout$attrs
```

```{r}
vg_render(spec)
```

## Loading data from a file

`vg_data()` can also point at a file instead of an inline data frame -- the
fictional `data/stocks.parquet` from the original design doc, made real as
a small `.csv` file shipped alongside this vignette. A relative path like
this resolves against wherever the rendered document itself ends up, the
same way a relative link or image would in an ordinary web page.

(Named `"aapl_file"` here, not `"aapl"` -- every `vg_render()` call on this
page shares one DuckDB instance behind the scenes, so two different specs
loading different data under the same name would collide.)

```{r}
spec1b <-
  vg_create() |>
  vg_data(name = "aapl_file", file = "data/stocks.csv", where = "Symbol = 'AAPL'") |>
  vg_line_y(data_from = "aapl_file", x = ~Date, y = ~Close) |>
  vg_attributes(width = 680, height = 200)

vg_render(spec1b)
```

## Combining marks and an interactor on one plot

Piping one mark straight into another appends it to the *same* plot,
without needing to wrap them in `vg_plot()` explicitly. Interactors
(`vg_interval_x()` here) can be piped in the same way.

```{r}
walk <- data.frame(t = 1:100, v = cumsum(rnorm(100)))

spec2 <-
  vg_create() |>
  vg_data(name = "walk", data = walk) |>
  vg_area_y(data_from = "walk", x = ~t, y = ~v, fill = "steelblue") |>
  vg_interval_x(as = param(brush)) |>
  vg_plot(width = 680, height = 200, name = "focus")

spec2
```

`param(brush)` is how a `$brush`-style reference to a Param/Selection gets
written in valid R -- `$brush` alone isn't parseable:

```{r}
param(brush)
```

```{r}
vg_render(spec2)
```

## Two linked plots with vg_vconcat()

This reproduces mosaic's own `overview-detail.yaml` example: an "overview"
plot with a brush interactor, stacked above a "detail" plot that's filtered
by whatever the brush selects. Drag a selection on the top plot to see the
bottom one update.

Passing separate mark/interactor chains to `vg_vconcat()` stacks them as
*separate* plots -- unlike piping them together, which (as above) keeps them
on one shared plot.

```{r}
overview <-
  vg_area_y(data_from = "walk", x = ~t, y = ~v, fill = "steelblue") |>
  vg_interval_x(as = param(brush)) |>
  vg_plot(width = 680, height = 150)

detail <-
  vg_area_y(
    data_from = "walk",
    filter_by = param(brush),
    x = ~t,
    y = ~v,
    fill = "steelblue"
  ) |>
  vg_plot(width = 680, height = 150)

spec3 <-
  vg_create() |>
  vg_data(name = "walk", data = walk) |>
  vg_vconcat(overview, detail)

spec3
```

```{r}
vg_render(spec3)
```

## Transform functions: binning and aggregation

Encoding formulas can use transform functions like `vg_bin()`/`vg_count()`
(mirroring mosaic's own `bin`/`count` transforms) instead of a bare column
reference. This reproduces the histogram pattern from mosaic's
`flights-200k.yaml` example (`x: {bin: delay}`, `y: {count:}`), binning a
continuous variable and counting rows per bin -- all computed by DuckDB, not
in R.

```{r}
delays <- data.frame(delay = rgamma(2000, shape = 2, rate = 0.3) - 5)

spec4 <-
  vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_attributes(width = 680, height = 200)

spec4
```

Every transform function also works called directly, without a formula --
useful for inspecting what it builds:

```{r}
vg_bin(delay, step = 2)
```

```{r}
vg_render(spec4)
```

## Generated marks and attributes

vgplotr now has a wrapper for every mark type mosaic supports (65, generated
from mosaic's own JSON schema by `data-raw/update-schema.R` -- see
`vg_hexbin()`, `vg_text()`, `vg_rule_x()`, ... alongside the marks used
above), and correctly recognizes every plot-level attribute (215) as
belonging to the plot, not the mark. That fixed a real bug along the way:
`marginLeft`/`marginRight`/`marginTop`/`marginBottom` and `aspectRatio` used
to only be checked for under invented names (`margin_left`, `aspect_ratio`)
that were never mosaic-spec's actual attribute names, so setting them
silently did nothing.

```{r}
spec4b <-
  vg_create() |>
  vg_data(name = "delays", data = delays) |>
  vg_rect_y(
    data_from = "delays",
    x = ~ vg_bin(delay, step = 2),
    y = ~ vg_count(),
    fill = "steelblue",
    inset = 0.5
  ) |>
  vg_rule_x(x = 0, stroke = "firebrick", strokeWidth = 2) |>
  vg_plot(width = 680, height = 200, marginLeft = 60, aspectRatio = 1)

vg_render(spec4b)
```

## Legends

A legend embedded in a plot (piped in alongside its marks, with no
`for_plot`) picks up a matching encoding automatically -- here, the color
mapping from `fill = ~group`.

```{r}
pts <- data.frame(
  x = rnorm(90),
  y = rnorm(90),
  group = sample(c("a", "b", "c"), 90, replace = TRUE)
)

spec5 <-
  vg_create() |>
  vg_data(name = "pts", data = pts) |>
  vg_dot(data_from = "pts", x = ~x, y = ~y, fill = ~group) |>
  vg_legend(type = "color") |>
  vg_plot(width = 400, height = 300)

vg_render(spec5)
```

## Layout-level inputs

Inputs like `vg_menu()`/`vg_slider()` are standalone widgets that live
alongside plots in the layout (via `vg_vconcat()`/`vg_hconcat()`), rather
than inside one. Here a menu drives a `param()` that's used directly as a
mark's `fill` (no formula needed, since it isn't a column mapping) -- pick a
different color and the plot updates.

```{r}
color_picker <- vg_menu(label = "Color", as = param(dotcolor), options = c("steelblue", "orange", "firebrick"))

colored_plot <-
  vg_dot(data_from = "pts", x = ~x, y = ~y, fill = param(dotcolor)) |>
  vg_plot(width = 400, height = 300)

spec6 <-
  vg_create() |>
  vg_data(name = "pts", data = pts) |>
  vg_vconcat(color_picker, colored_plot)

vg_render(spec6)
```

## What happens when two marks disagree

Plot-level attributes (like `width`) accumulate silently across marks when
they agree, but a genuine conflict is not resolved silently -- the later
value wins, and you get a warning so it doesn't happen by surprise.

```{r}
plot3 <-
  vg_dot(x = ~a, y = ~b, width = 400) |>
  vg_line_y(x = ~a, y = ~c, width = 680)

plot3$attrs$width
```

## `sql()`/`agg()`: raw SQL expressions

`sql()` (and `agg()`, for expressions with an aggregate function) wraps a
raw SQL expression for encodings that transforms and plain column
references can't express -- mosaic-spec's own `{sql: "..."}`/`{agg: "..."}`.
This finally realizes the very first example from the original design doc
(`design/api-brainstorming.qmd`): a slider-controlled "bias" parameter
added to a plotted value, computed live by DuckDB. A `param()` reference
can be written directly in the SQL text (`sql("v + $bias")`) or built up
from pieces, as below -- both are equivalent.

```{r}
bias_picker <- vg_slider(label = "Bias", as = param(bias), min = -20, max = 20, step = 1, value = 0)

biased_plot <-
  vg_area_y(data_from = "walk", x = ~t, y = sql("v + ", param(bias)), fill = "steelblue") |>
  vg_plot(width = 680, height = 200, yDomain = c(-30, 30))

spec7 <-
  vg_create() |>
  vg_data(name = "walk", data = walk) |>
  vg_vconcat(bias_picker, biased_plot)

vg_render(spec7)
```
