---
title: "Cross-Filter Flights (10M)"
subtitle: "Histograms over 10 million flights, with automatic pre-aggregation for efficient cross-filtering."
format: html
engine: knitr
---

Recreates [Cross-Filter Flights
(10M)](https://idl.uw.edu/mosaic/examples/flights-10m.html) from the
mosaic/vgplot example gallery. Once loaded, automatic pre-aggregation
optimizations enable efficient cross-filtered selections.

Note: the underlying `flights-10m.parquet` file is ~10 million rows,
hosted outside mosaic's usual `data/` directory (see "Linear
Regression 10M" in the Data Transformation batch for the same data
source). Rendering this one may be slow.

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)

flights_url <- "https://pub-1da360b43ceb401c809f68ca37c7f8a4.r2.dev/data/flights-10m.parquet"

delay_plot <-
  vg_mark_rect_y(
    data_from = "flights10m", filter_by = brush,
    x = ~ vg_bin(delay), y = ~ vg_count(), fill = "steelblue",
    inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", x_label = "Arrival Delay (min)", y_tick_format = "s", width = 600, height = 200)

time_plot <-
  vg_mark_rect_y(
    data_from = "flights10m", filter_by = brush,
    x = ~ vg_bin(time), y = ~ vg_count(), fill = "steelblue",
    inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", x_label = "Departure Time (hour)", y_tick_format = "s", width = 600, height = 200)

distance_plot <-
  vg_mark_rect_y(
    data_from = "flights10m", filter_by = brush,
    x = ~ vg_bin(distance), y = ~ vg_count(), fill = "steelblue",
    inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", x_label = "Flight Distance (miles)", y_tick_format = "s", width = 600, height = 200)

spec <-
  vg_create() |>
  vg_meta(
    title = "Cross-Filter Flights (10M)",
    description = paste(
      "Histograms showing arrival delay, departure time, and distance flown for 10 million",
      "flights.\nOnce loaded, automatic pre-aggregation optimizations enable efficient",
      "cross-filtered selections."
    )
  ) |>
  vg_data(name = "flights10m", query = sprintf(
    "SELECT GREATEST(-60, LEAST(ARR_DELAY, 180))::DOUBLE AS delay, DISTANCE AS distance, DEP_TIME AS time FROM %s",
    shQuote(flights_url, type = "sh")
  )) |>
  vg_params(brush = list(select = "crossfilter")) |>
  vg_vconcat(delay_plot, time_plot, distance_plot)

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/flights-10m.html) exactly,
aside from object key order.
:::
