---
title: "Cross-Filter Flights (200k)"
subtitle: "Histograms of arrival delay, departure time, and flight distance, cross-filtered by brushing."
format: html
engine: knitr
---

Recreates [Cross-Filter Flights
(200k)](https://idl.uw.edu/mosaic/examples/flights-200k.html) from the
mosaic/vgplot example gallery. Select a histogram region to
cross-filter the charts. Each plot uses an `intervalX` interactor to
populate a shared Selection with `crossfilter` resolution.

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)

delay_plot <-
  vg_mark_rect_y(
    data_from = "flights", filter_by = brush,
    x = ~ vg_bin(delay), y = ~ vg_count(), fill = "steelblue",
    inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", x_label = "Arrival Delay (min)", y_tick_format = "s", width = 600, height = 200)

time_plot <-
  vg_mark_rect_y(
    data_from = "flights", filter_by = brush,
    x = ~ vg_bin(time), y = ~ vg_count(), fill = "steelblue",
    inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", x_label = "Departure Time (hour)", y_tick_format = "s", width = 600, height = 200)

distance_plot <-
  vg_mark_rect_y(
    data_from = "flights", filter_by = brush,
    x = ~ vg_bin(distance), y = ~ vg_count(), fill = "steelblue",
    inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", x_label = "Flight Distance (miles)", y_tick_format = "s", width = 600, height = 200)

spec <-
  vg_create() |>
  vg_meta(
    title = "Cross-Filter Flights (200k)",
    description = paste(
      "Histograms showing arrival delay, departure time, and distance flown for over",
      "200,000 flights. Select a histogram region to cross-filter the charts. Each plot",
      "uses an `intervalX` interactor to populate a shared Selection with `crossfilter`",
      "resolution."
    )
  ) |>
  vg_data(name = "flights", file = vg_data_url("flights-200k.parquet")) |>
  vg_params(brush = list(select = "crossfilter")) |>
  vg_vconcat(delay_plot, time_plot, distance_plot)

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/flights-200k.html) exactly,
aside from `data.flights.file` (`vg_data_url()` vs. a relative path)
and object key order.
:::
