---
title: "Pan & Zoom"
subtitle: "Linked panning and zooming across four scatter plots sharing axes."
format: html
engine: knitr
---

Recreates [Pan & Zoom](https://idl.uw.edu/mosaic/examples/pan-zoom.html)
from the mosaic/vgplot example gallery. Drag to pan, scroll to zoom.
`panZoom` interactors update a set of bound selections, one per unique
axis.

Like "Linear Regression"'s brush, the four axis selections (`xs`/`ys`/
`zs`/`ws`) are each created implicitly by being the target of a
`panZoom(x =, y =)` call -- no `vg_params()` declaration needed, since
each is used only within the plots that reference it.

```{r setup}
library(vgplotr)
```

```{r}
xs <- param(xs)
ys <- param(ys)
zs <- param(zs)
ws <- param(ws)

plot1 <-
  vg_mark_frame() |>
  vg_mark_dot(data_from = "penguins", x = ~bill_length, y = ~bill_depth, fill = ~species, r = 2, clip = TRUE) |>
  vg_pan_zoom(x = xs, y = ys) |>
  vg_plot(width = 320, height = 240)

plot2 <-
  vg_mark_frame() |>
  vg_mark_dot(data_from = "penguins", x = ~bill_length, y = ~flipper_length, fill = ~species, r = 2, clip = TRUE) |>
  vg_pan_zoom(x = xs, y = zs) |>
  vg_plot(width = 320, height = 240)

plot3 <-
  vg_mark_frame() |>
  vg_mark_dot(data_from = "penguins", x = ~body_mass, y = ~bill_depth, fill = ~species, r = 2, clip = TRUE) |>
  vg_pan_zoom(x = ws, y = ys) |>
  vg_plot(width = 320, height = 240)

plot4 <-
  vg_mark_frame() |>
  vg_mark_dot(data_from = "penguins", x = ~body_mass, y = ~flipper_length, fill = ~species, r = 2, clip = TRUE) |>
  vg_pan_zoom(x = ws, y = zs) |>
  vg_plot(width = 320, height = 240)

spec <-
  vg_create() |>
  vg_meta(
    title = "Pan & Zoom",
    description = paste(
      "Linked panning and zooming across plots: drag to pan, scroll to zoom.",
      "`panZoom` interactors update a set of bound selections, one per unique axis."
    )
  ) |>
  vg_data(name = "penguins", file = vg_data_url("penguins.parquet")) |>
  vg_hconcat(
    vg_vconcat(plot1, vg_vspace(10), plot2),
    vg_hspace(10),
    vg_vconcat(plot3, vg_vspace(10), plot4)
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/pan-zoom.html) exactly, aside
from `data.penguins.file` (`vg_data_url()` vs. a relative path) and
object key order.
:::
