---
title: "NYC Taxi Rides"
subtitle: "Pickup and dropoff points for 1M NYC taxi rides, with linked brushing across all three plots."
format: html
engine: knitr
---

Recreates [NYC Taxi
Rides](https://idl.uw.edu/mosaic/examples/nyc-taxi-rides.html) from
the mosaic/vgplot example gallery. This example projects lon/lat
coordinates in the database upon load. Select a region in one plot to
filter the other. What spatial patterns can you find? Requires the
DuckDB `spatial` extension.

`vg_config(extensions = "spatial")` loads the DuckDB extension mosaic's
`ST_Transform()`/`ST_Point()`/etc. spatial functions need -- mosaic-spec's
top-level `config` object, distinct from `vg_meta()`'s inert descriptive
metadata. All three plots (pickups, dropoffs, and the pickup-hour
histogram) share one `crossfilter` Selection (`filter`), each
contributing its own `intervalX`/`intervalXY` brush to it.

```{r setup}
library(vgplotr)
```

```{r}
filter_p <- param(filter)

rides_url <- "https://pub-1da360b43ceb401c809f68ca37c7f8a4.r2.dev/data/nyc-rides-2010.parquet"

pickup_plot <-
  vg_mark_raster(data_from = "trips", filter_by = filter_p, x = ~px, y = ~py, bandwidth = 0) |>
  vg_interval_xy(as = filter_p) |>
  vg_mark_text(
    data_from = list(list(label = "Taxi Pickups")),
    dx = 10, dy = 10, text = ~label, fill = "black", font_size = "1.2em", frame_anchor = "top-left"
  ) |>
  vg_plot(
    width = 335, height = 550, margin = 0, x_axis = NULL, y_axis = NULL,
    x_domain = c(975000, 1005000), y_domain = c(190000, 240000),
    color_scale = "symlog", color_scheme = "blues"
  )

dropoff_plot <-
  vg_mark_raster(data_from = "trips", filter_by = filter_p, x = ~dx, y = ~dy, bandwidth = 0) |>
  vg_interval_xy(as = filter_p) |>
  vg_mark_text(
    data_from = list(list(label = "Taxi Dropoffs")),
    dx = 10, dy = 10, text = ~label, fill = "black", font_size = "1.2em", frame_anchor = "top-left"
  ) |>
  vg_plot(
    width = 335, height = 550, margin = 0, x_axis = NULL, y_axis = NULL,
    x_domain = c(975000, 1005000), y_domain = c(190000, 240000),
    color_scale = "symlog", color_scheme = "oranges"
  )

spec <-
  vg_create() |>
  vg_meta(
    title = "NYC Taxi Rides",
    description = paste(
      "Pickup and dropoff points for 1M NYC taxi rides on Jan 1-3, 2010.",
      "This example projects lon/lat coordinates in the database upon load.",
      "Select a region in one plot to filter the other.",
      "What spatial patterns can you find?",
      "Requires the DuckDB `spatial` extension.",
      sep = "\n"
    )
  ) |>
  vg_config(extensions = "spatial") |>
  vg_data(
    name = "rides", file = rides_url,
    select = c(
      "pickup_datetime::TIMESTAMP AS datetime",
      "ST_Transform(ST_Point(pickup_latitude, pickup_longitude), 'EPSG:4326', 'ESRI:102718') AS pick",
      "ST_Transform(ST_Point(dropoff_latitude, dropoff_longitude), 'EPSG:4326', 'ESRI:102718') AS drop"
    )
  ) |>
  vg_data(
    name = "trips",
    query = paste0(
      "SELECT\n  (HOUR(datetime) + MINUTE(datetime)/60) AS time,\n  ST_X(pick) AS px, ",
      "ST_Y(pick) AS py,\n  ST_X(drop) AS dx, ST_Y(drop) AS dy\nFROM rides\n"
    )
  ) |>
  vg_params(filter = list(select = "crossfilter")) |>
  vg_vconcat(
    vg_hconcat(pickup_plot, vg_hspace(10), dropoff_plot),
    vg_vspace(10),
    vg_mark_rect_y(data_from = "trips", x = ~ vg_bin(time), y = ~ vg_count(), fill = "steelblue", inset = 0.5) |>
      vg_interval_x(as = filter_p) |>
      vg_plot(y_tick_format = "s", x_label = "Pickup Hour →", width = 680, height = 100)
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/nyc-taxi-rides.html) exactly
(this example's data is already a fully-qualified external URL, not
under mosaic's own `data/` directory, so no `vg_data_url()` difference
here), aside from object key order.
:::
