---
title: "Two Globes"
subtitle: "Two rotatable globes overlaid"
format: html
engine: knitr
---

Based on [Earthquakes
Globe](https://idl.uw.edu/mosaic/examples/earthquakes-globe.html) from
the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-earthquake-globe)).
To show land masses, this example loads and parses TopoJSON data in
the database. Requires the DuckDB `spatial` extension.

The `rotate` param combines two other params (`longitude`/`latitude`,
each driven by its own slider) into one value -- mosaic's own
`rotate: [$longitude, $latitude]` pattern -- passed straight through as
`vg_params(rotate = list(lon, lat))`.

```{r setup}
library(vgplotr)
```

```{r}
lon1 <- param(longitude1)
lat1 <- param(latitude1)
lon2 <- param(longitude2)
lat2 <- param(latitude2)
rotate_p <- param(rotate1)
rotate_q <- param(rotate2)

spec <-
  vg_create() |>
  vg_meta(
    title = "Two Globes",
    description = paste(
      "A rotatable globe of earthquake activity. To show land masses, this example loads",
      "and parses TopoJSON data in the database. Requires the DuckDB `spatial` extension."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-earthquake-globe)."
  ) |>
  vg_data(
    name = "land",
    type = "spatial",
    file = vg_data_url("countries-110m.json"),
    layer = "land"
  ) |>
  vg_params(longitude = -180, latitude = -30, rotate1 = list(lon1, lat1)) |>
  vg_params(longitude = -180, latitude = -30, rotate2 = list(lon2, lat2)) |>
  vg_vconcat(
    vg_hconcat(
      vg_slider(
        label = "Longitude",
        as = lon1,
        min = -180,
        max = 180,
        step = 1
      ),
      vg_slider(label = "Latitude", as = lat1, min = -90, max = 90, step = 1)
    ),
    vg_hconcat(
      vg_slider(
        label = "Longitude",
        as = lon2,
        min = -180,
        max = 180,
        step = 1
      ),
      vg_slider(label = "Latitude", as = lat2, min = -90, max = 90, step = 1)
    ),
    vg_mark_geo(
      data_from = "land",
      geometry = ~ vg_geojson(geom),
      fill = "red",
      fill_opacity = 0.2
    ) |>
      vg_mark_sphere(
        projection_rotate = rotate_p,
      ) |>
      vg_mark_geo(
        data_from = "land",
        geometry = ~ vg_geojson(geom),
        fill = "steelblue",
        fill_opacity = 0.2
      ) |>
      vg_mark_sphere(
        projection_rotate = rotate_q
      ) |>
      vg_plot(
        margin = 10,
        style = "overflow: visible;",
        projection_type = "orthographic"
      )
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## JSON 

```{r}
cat(to_json(spec))
```

:::
