---
title: "Symbol Plots"
subtitle: "Two scatter plots with dot marks: one with stroked symbols, the other filled. Drop-down menus control which data table columns are plotted."
format: html
engine: knitr
---

Recreates [Symbol Plots](https://idl.uw.edu/mosaic/examples/symbol-plots.html)
from the mosaic/vgplot example gallery. Demonstrates dynamic column
selection: `x =`/`y =` are bound to `vg_column(param(...))`, so the menu
inputs change which data column each axis encodes at runtime.

```{r setup}
library(vgplotr)
```

```{r}
x_param <- param(x)
y_param <- param(y)
columns <- c("body_mass", "flipper_length", "bill_depth", "bill_length")

menus <- vg_hconcat(
  vg_menu(label = "Y", options = columns, as = y_param),
  vg_menu(label = "X", options = columns, as = x_param)
)

stroked <-
  vg_mark_dot(
    data_from = "penguins",
    x = ~ vg_column(x_param),
    y = ~ vg_column(y_param),
    stroke = "species",
    symbol = "species"
  ) |>
  vg_guide_all(grid = TRUE) |>
  vg_guide_x(label = "Body mass (g) →") |>
  vg_guide_y(label = "↑ Flipper length (mm)") |>
  vg_plot(name = "stroked")

filled <-
  vg_mark_dot(
    data_from = "penguins",
    x = ~ vg_column(x_param),
    y = ~ vg_column(y_param),
    fill = "species",
    symbol = "species"
  ) |>
  vg_guide_all(grid = TRUE) |>
  vg_guide_x(label = "Body mass (g) →") |>
  vg_guide_y(label = "↑ Flipper length (mm)") |>
  vg_plot(name = "filled")

spec <-
  vg_create() |>
  vg_meta(
    title = "Symbol Plots",
    description = "Two scatter plots with `dot` marks: one with stroked symbols, the other filled. Drop-down menus control which data table columns are plotted."
  ) |>
  vg_data(name = "penguins", file = vg_data_url("penguins.parquet")) |>
  vg_params(x = "body_mass", y = "flipper_length") |>
  vg_vconcat(
    menus,
    vg_vspace(10),
    vg_hconcat(stroked, vg_legend_symbol(for_plot = "stroked", columns = 1)),
    vg_vspace(20),
    vg_hconcat(filled, vg_legend_symbol(for_plot = "filled", columns = 1))
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/symbol-plots.html) apart from
`data.penguins.file` (a full URL from `vg_data_url()` here, vs. mosaic's
own relative `data/penguins.parquet` path) and object key order.
</details>
