---
title: "Voronoi Diagram"
subtitle: "The voronoi mark shows the regions closest to each point."
format: html
engine: knitr
---

Recreates [Voronoi Diagram](https://idl.uw.edu/mosaic/examples/voronoi.html)
from the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-voronoi-scatterplot)).

The `voronoi` mark shows the regions closest to each point. It is
[constructed from its
dual](https://observablehq.com/@mbostock/the-delaunays-dual), a Delaunay
triangle mesh. Reveal triangulations or convex hulls using the dropdowns.

```{r setup}
library(vgplotr)
```

```{r}
mesh <- param(mesh)
hull <- param(hull)

spec <-
  vg_create() |>
  vg_meta(
    title = "Voronoi Diagram",
    description = paste(
      "The `voronoi` mark shows the regions closest to each point. It is",
      "[constructed from its dual](https://observablehq.com/@mbostock/the-delaunays-dual),",
      "a Delaunay triangle mesh. Reveal triangulations or convex hulls using the dropdowns."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-voronoi-scatterplot)."
  ) |>
  vg_data(name = "penguins", file = vg_data_url("penguins.parquet")) |>
  vg_params(mesh = 0, hull = 0) |>
  vg_vconcat(
    vg_mark_voronoi(
      data_from = "penguins",
      x = ~bill_length,
      y = ~bill_depth,
      stroke = "white",
      stroke_width = 1,
      stroke_opacity = 0.5,
      fill = ~species,
      fill_opacity = 0.2
    ) |>
      vg_mark_hull(
        data_from = "penguins",
        x = ~bill_length,
        y = ~bill_depth,
        stroke = ~species,
        stroke_opacity = hull,
        stroke_width = 1.5
      ) |>
      vg_mark_delaunay_mesh(
        data_from = "penguins",
        x = ~bill_length,
        y = ~bill_depth,
        z = ~species,
        stroke = ~species,
        stroke_opacity = mesh,
        stroke_width = 1
      ) |>
      vg_mark_dot(
        data_from = "penguins",
        x = ~bill_length,
        y = ~bill_depth,
        fill = ~species,
        r = 2
      ) |>
      vg_mark_frame() |>
      vg_plot(inset = 10, width = 680),
    vg_hconcat(
      vg_menu(
        label = "Delaunay Mesh",
        options = list(
          list(value = 0, label = "Hide"),
          list(value = 0.5, label = "Show")
        ),
        as = mesh
      ),
      vg_hspace(5),
      vg_menu(
        label = "Convex Hull",
        options = list(
          list(value = 0, label = "Hide"),
          list(value = 1, label = "Show")
        ),
        as = hull
      )
    )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/voronoi.html) exactly, aside
from `data.penguins.file` (`vg_data_url()` vs. a relative path) and
object key order.
</details>
