---
title: "Airline Travelers"
subtitle: "A labeled line chart comparing airport travelers in 2019 and 2020."
format: html
engine: knitr
---

Recreates [Airline
Travelers](https://idl.uw.edu/mosaic/examples/airline-travelers.html) from
the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-labeled-line-chart)).

Demonstrates a `query`-only data source (a derived temporary table
computed with a raw SQL query, no file behind it) and a literal,
per-mark constant text label -- `text = list("2019")`, not a formula --
which is mosaic's `["2019"]` array-literal-channel convention (an array
forces a literal value; a bare string would be interpreted as a column
name instead).

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "Airline Travelers",
    description = "A labeled line chart comparing airport travelers in 2019 and 2020.",
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-labeled-line-chart)."
  ) |>
  vg_data(name = "travelers", file = vg_data_url("travelers.parquet")) |>
  vg_data(
    name = "endpoint",
    query = "SELECT * FROM travelers ORDER BY date DESC LIMIT 1\n"
  ) |>
  vg_mark_rule_y(data_from = c(0)) |>
  vg_mark_line_y(
    data_from = "travelers",
    x = ~date,
    y = ~previous,
    stroke_opacity = 0.35
  ) |>
  vg_mark_line_y(data_from = "travelers", x = ~date, y = ~current) |>
  vg_mark_text(
    data_from = "endpoint",
    x = ~date,
    y = ~previous,
    text = list("2019"),
    fill_opacity = 0.5,
    line_anchor = "bottom",
    dy = -6
  ) |>
  vg_mark_text(
    data_from = "endpoint",
    x = ~date,
    y = ~current,
    text = list("2020"),
    line_anchor = "top",
    dy = 6
  ) |>
  vg_guide_y(grid = TRUE, label = "↑ Travelers per day", tick_format = "s")

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/airline-travelers.html) exactly,
aside from `data.travelers.file` (a full URL from `vg_data_url()` vs.
mosaic's relative `data/travelers.parquet`) and object key order.
</details>
