---
title: "Sorted Bars"
subtitle: "Sort and limit an aggregate bar chart of gold medals by country."
format: html
engine: knitr
---

Recreates [Sorted Bars](https://idl.uw.edu/mosaic/examples/sorted-bars.html)
from the mosaic/vgplot example gallery.

The menu's `as = query` output is a mosaic **Selection**
(`vg.Selection.intersect()` on the JS side), not a plain Param -- declared
in `vg_params()` as a list with a `select` field (mosaic-spec's `Selection`
shape, `{"select": "intersect"}`) rather than a bare value. The bar mark
then reads `filter_by = query` to filter by the menu's current selection.

```{r setup}
library(vgplotr)
```

```{r}
query <- param(query)

spec <-
  vg_create() |>
  vg_meta(
    title = "Sorted Bars",
    description = "Sort and limit an aggregate bar chart of gold medals by country."
  ) |>
  vg_data(name = "athletes", file = vg_data_url("athletes.parquet")) |>
  vg_params(query = list(select = "intersect")) |>
  vg_vconcat(
    vg_menu(
      label = "Sport",
      as = query,
      from = "athletes",
      column = "sport",
      value = "aquatics"
    ),
    vg_vspace(10),
    vg_mark_bar_x(
      data_from = "athletes",
      filter_by = query,
      x = ~ vg_sum(gold),
      y = ~nationality,
      fill = "steelblue",
      sort = list(y = "-x", limit = 10)
    ) |>
      vg_plot(
        x_label = "Gold Medals",
        y_label = "Nationality",
        y_label_anchor = "top",
        margin_top = 15
      )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/sorted-bars.html) exactly, aside
from `data.athletes.file` (`vg_data_url()` vs. a relative path) and object
key order.
</details>
