---
title: "Seattle Weather Pivot"
subtitle: "A DuckDB PIVOT query reshapes daily weather observations into a cross-tab, shown in a sortable table."
format: html
engine: knitr
---

Recreates [Seattle Weather
Pivot](https://idl.uw.edu/mosaic/examples/seattle-weather-pivot.html)
from the mosaic/vgplot example gallery. A DuckDB `PIVOT` query reshapes
Seattle's daily weather observations into a cross-tab: one row per
year, with a column counting the days of each weather type. The
pivoted result is shown in a sortable `table` view. Click a column
header to sort.

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "Seattle Weather Pivot",
    description = paste(
      "A DuckDB `PIVOT` query reshapes Seattle's daily weather observations into a",
      "cross-tab: one row per year, with a column counting the days of each weather type.",
      "The pivoted result is shown in a sortable `table` view. Click a column header to sort."
    )
  ) |>
  vg_data(name = "seattle_weather", file = vg_data_url("seattle-weather.parquet")) |>
  vg_data(
    name = "weatherByYear",
    query = paste0(
      "PIVOT (SELECT *, year(date) AS year FROM seattle_weather) ON weather IN ",
      "('drizzle', 'fog', 'rain', 'snow', 'sun') USING count(*) GROUP BY year ORDER BY year\n"
    )
  ) |>
  vg_vconcat(vg_table(from = "weatherByYear", align = list(year = "left"), width = list(year = 80), height = 180))

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Mosaic's own JSON spec puts the `table` input directly at the top level
(no `vconcat` wrapper), since it's the only thing in the layout -- same
cosmetic difference as "Sortable Table" (see `NOTES.md`). Aside from
that and `data.seattle_weather.file` (`vg_data_url()` vs. a relative
path), this matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/seattle-weather-pivot.html).
</details>
