---
title: "Athlete Height Intervals"
subtitle: "Confidence intervals of Olympic athlete heights, batched into groups of 10 samples per sport."
format: html
engine: knitr
---

Recreates [Athlete Height
Intervals](https://idl.uw.edu/mosaic/examples/athlete-height.html) from
the mosaic/vgplot example gallery. Use the samples slider to see how the
intervals update as the sample size increases (as in online
aggregation). For each sport, the numbers on the right show the maximum
number of athletes in the full dataset.

```{r setup}
library(vgplotr)
```

```{r}
ci <- param(ci)
query <- param(query)

spec <-
  vg_create() |>
  vg_meta(
    title = "Athlete Height Intervals",
    description = paste(
      "Confidence intervals of Olympic athlete heights, in meters. Data are batched into",
      "groups of 10 samples per sport. Use the samples slider to see how the intervals",
      "update as the sample size increases (as in online aggregation). For each sport, the",
      "numbers on the right show the maximum number of athletes in the full dataset."
    )
  ) |>
  vg_data(
    name = "athletesBatched",
    file = vg_example_url("athletes.parquet"),
    select = c(
      "*",
      "10 * CEIL(ROW_NUMBER() OVER (PARTITION BY sport) / 10) AS batch"
    ),
    where = "height IS NOT NULL"
  ) |>
  vg_params(ci = 0.95, query = list(select = "single")) |>
  vg_hconcat(
    vg_vconcat(
      vg_hconcat(
        vg_slider(
          select = "interval",
          as = query,
          column = "batch",
          from = "athletesBatched",
          step = 10,
          value = 20,
          label = "Max Samples"
        ),
        vg_slider(
          as = ci,
          min = 0.5,
          max = 0.999,
          step = 0.001,
          label = "Conf. Level"
        )
      ),
      vg_mark_errorbar_x(
        data_from = "athletesBatched",
        filter_by = query,
        ci = ci,
        x = ~height,
        y = ~sport,
        stroke = ~sex,
        stroke_width = 1,
        marker = "tick",
        sort = list(y = "-x")
      ) |>
        vg_mark_text(
          data_from = "athletesBatched",
          frame_anchor = "right",
          font_size = 8,
          fill = "#999",
          dx = 25,
          text = ~ vg_count(),
          y = ~sport
        ) |>
        vg_plot(
          name = "heights",
          x_domain = c(1.5, 2.1),
          y_domain = "Fixed",
          y_grid = TRUE,
          y_label = NULL,
          margin_top = 5,
          margin_left = 105,
          margin_right = 30,
          height = 420
        ),
      vg_legend_color(for_plot = "heights")
    )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/athlete-height.html) exactly,
aside from `data.athletesBatched.file` (`vg_example_url()` vs. a relative
path) and object key order.
</details>
