---
title: "Linear Regression 10M"
subtitle: "A linear regression plot over 10 million flight records, with optimized selection updates using materialized sample views."
format: html
engine: knitr
---

Recreates [Linear Regression
10M](https://idl.uw.edu/mosaic/examples/linear-regression-10m.html)
from the mosaic/vgplot example gallery. Regression computation is
performed in the database, with optimized selection updates using
pre-aggregated materialized views. The area around a regression line
shows a 95% confidence interval. Select a region to view regression
results for a data subset.

Note: the underlying `flights-10m.parquet` file is ~10 million rows,
hosted outside mosaic's usual `data/` directory (a Cloudflare R2
bucket, referenced directly inside the SQL query text below rather than
via `vg_data_url()`, which only covers files under mosaic's own GitHub
`data/` directory). Rendering this one may be slow.

```{r setup}
library(vgplotr)
```

```{r}
data_p <- param(data)
query <- param(query)

flights_url <- "https://pub-1da360b43ceb401c809f68ca37c7f8a4.r2.dev/data/flights-10m.parquet"

spec <-
  vg_create() |>
  vg_meta(
    title = "Linear Regression 10M",
    description = paste(
      "A linear regression plot predicting flight arrival delay based on the time of",
      "departure, over 10 million flight records. Regression computation is performed in",
      "the database, with optimized selection updates using pre-aggregated materialized",
      "views. The area around a regression line shows a 95% confidence interval. Select a",
      "region to view regression results for a data subset."
    )
  ) |>
  vg_data(name = "flights10m", query = sprintf(
    "SELECT GREATEST(-60, LEAST(ARR_DELAY, 180))::DOUBLE AS delay, DISTANCE AS distance, DEP_TIME AS time FROM %s",
    shQuote(flights_url, type = "sh")
  )) |>
  vg_data(name = "flights10p", query = "SELECT * FROM flights10m USING SAMPLE 10%") |>
  vg_data(name = "flights5p", query = "SELECT * FROM flights10m USING SAMPLE 5%") |>
  vg_data(name = "flights1p", query = "SELECT * FROM flights10m USING SAMPLE 1%") |>
  vg_params(data = "flights10m") |>
  vg_vconcat(
    vg_menu(
      label = "Sample", as = data_p,
      options = list(
        list(value = "flights10m", label = "Full Data"),
        list(value = "flights10p", label = "10% Sample"),
        list(value = "flights5p", label = "5% Sample"),
        list(value = "flights1p", label = "1% Sample")
      )
    ),
    vg_vspace(10),
    vg_mark_raster(
      data_from = data_p, x = ~time, y = ~delay,
      pixel_size = 4, pad = 0, image_rendering = "pixelated"
    ) |>
      vg_mark_regression_y(data_from = data_p, x = ~time, y = ~delay, stroke = "gray") |>
      vg_mark_regression_y(data_from = data_p, filter_by = query, x = ~time, y = ~delay, stroke = "firebrick") |>
      vg_interval_xy(as = query, brush = list(fillOpacity = 0, stroke = "currentColor")) |>
      vg_plot(
        x_domain = c(0, 24), y_domain = c(-60, 180),
        color_scale = "symlog", color_scheme = "blues", color_domain = "Fixed"
      )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/linear-regression-10m.html)
exactly, aside from object key order.
</details>
