---
title: "Overview + Detail"
subtitle: "Select the top overview series to zoom the focus view below."
format: html
engine: knitr
---

Recreates [Overview +
Detail](https://idl.uw.edu/mosaic/examples/overview-detail.html) from
the mosaic/vgplot example gallery. An `intervalX` interactor updates a
selection that filters the focus view. The `line` and `area` marks can
apply M4 optimization to reduce the number of data points returned:
rather than draw all points, a dramatically smaller subset can still
faithfully represent these area charts.

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)

spec <-
  vg_create() |>
  vg_meta(
    title = "Overview + Detail",
    description = paste(
      'Select the top "overview" series to zoom the "focus" view below. An `intervalX`',
      "interactor updates a selection that filters the focus view. The `line` and `area`",
      "marks can apply M4 optimization to reduce the number of data points returned: rather",
      "than draw all points, a dramatically smaller subset can still faithfully represent",
      "these area charts."
    )
  ) |>
  vg_data(name = "walk", file = vg_example_url("random-walk.parquet")) |>
  vg_vconcat(
    vg_mark_area_y(data_from = "walk", x = ~t, y = ~v, fill = "steelblue") |>
      vg_interval_x(as = brush) |>
      vg_plot(width = 680, height = 200),
    vg_mark_area_y(data_from = "walk", filter_by = brush, x = ~t, y = ~v, fill = "steelblue") |>
      vg_plot(y_domain = "Fixed", width = 680, height = 200)
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/overview-detail.html) exactly,
aside from `data.walk.file` (`vg_example_url()` vs. a relative path) and
object key order.
</details>
