---
title: "Moving Average"
subtitle: "Daily reported COVID-19 cases in Berlin, smoothed with a moving average over various window frames."
format: html
engine: knitr
---

Recreates [Moving Average](https://idl.uw.edu/mosaic/examples/moving-average.html)
from the mosaic/vgplot example gallery. This plot shows daily reported
COVID-19 cases from March 3 (day 1) to May 5, 2020 (day 64) in Berlin,
Germany. We can smooth the raw counts using a moving average over
various choices of window query frames.

```{r setup}
library(vgplotr)
```

```{r}
frame <- param(frame)

spec <-
  vg_create() |>
  vg_meta(
    title = "Moving Average",
    description = paste(
      "This plot shows daily reported COVID-19 cases from March 3 (day 1) to May 5, 2020",
      "(day 64) in Berlin, Germany, as reported by the Robert Koch Institute. We can smooth",
      "the raw counts using a moving average over various choices of window query frames."
    ),
    credit = "Adapted from the [Arquero window query tutorial](https://observablehq.com/@uwdata/working-with-window-queries)."
  ) |>
  vg_data(name = "cases", file = vg_data_url("berlin-covid.parquet")) |>
  vg_params(frame = c(-6, 0)) |>
  vg_vconcat(
    vg_mark_rect_y(
      data_from = "cases", x1 = ~day, x2 = ~ sql("day + 1"), inset = 1, y = ~cases, fill = "steelblue"
    ) |>
      vg_mark_line_y(
        data_from = "cases", x = ~ sql("day + 0.5"),
        y = ~ vg_avg(cases, orderby = "day", rows = frame),
        curve = "monotone-x", stroke = "currentColor"
      ) |>
      vg_plot(x_label = "day", width = 680, height = 300),
    vg_menu(
      label = "Window Frame", as = frame,
      options = list(
        list(label = "7-day moving average, with prior 6 days: [-6, 0]", value = c(-6, 0)),
        list(label = "7-day moving average, centered at current day: [-3, 3]", value = c(-3, 3)),
        list(label = "Moving average, with all prior days [-∞, 0]", value = list(NULL, 0)),
        list(label = "Global average [-∞, +∞]", value = list(NULL, NULL))
      )
    )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/moving-average.html) exactly,
aside from `data.cases.file` (`vg_data_url()` vs. a relative path) and
object key order.
</details>
