---
title: "Line Density"
subtitle: "Density of stock price line series, normalized and unnormalized, with linked zoom."
format: html
engine: knitr
---

Recreates [Line Density](https://idl.uw.edu/mosaic/examples/line-density.html)
from the mosaic/vgplot example gallery. The `denseLine` mark shows the
densities of line series, here for a collection of stock prices. The
top plot normalizes by arc length to remove the vertical artifacts
visible in the unnormalized plot below. Select a region in the lower
plot to zoom the upper plot. The bandwidth slider smooths the data,
while the pixel size menu adjusts the raster resolution.

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)
bandwidth <- param(bandwidth)
pixel_size <- param(pixelSize)
scheme_color <- param(schemeColor)
scale_color <- param(scaleColor)

top_plot <-
  vg_mark_dense_line(
    data_from = "stocks_after_2006", filter_by = brush, x = ~Date, y = ~Close, z = ~Symbol, fill = "density",
    bandwidth = bandwidth, pixel_size = pixel_size
  ) |>
  vg_plot(
    color_scheme = scheme_color, color_scale = scale_color,
    y_label = "Close (Normalized) ↑", y_nice = TRUE,
    margins = list(left = 30, top = 20, right = 0), width = 680, height = 240
  )

bottom_plot <-
  vg_mark_dense_line(
    data_from = "stocks_after_2006", x = ~Date, y = ~Close, z = ~Symbol, fill = "density", normalize = FALSE,
    bandwidth = bandwidth, pixel_size = pixel_size
  ) |>
  vg_interval_xy(as = brush) |>
  vg_plot(
    color_scheme = scheme_color, color_scale = scale_color,
    y_label = "Close (Unnormalized) ↑", y_nice = TRUE,
    margins = list(left = 30, top = 20, right = 0), width = 680, height = 240
  )

spec <-
  vg_create() |>
  vg_meta(
    title = "Line Density",
    description = paste(
      "The `denseLine` mark shows the densities of line series, here for a collection of",
      "stock prices. The top plot normalizes by arc length to remove the vertical artifacts",
      "visible in the unnormalized plot below. Select a region in the lower plot to zoom the",
      "upper plot. The bandwidth slider smooths the data, while the pixel size menu adjusts",
      "the raster resolution."
    )
  ) |>
  vg_data(
    name = "stocks_after_2006", file = vg_data_url("stocks_after_2006.parquet"),
    select = c("Symbol", "Close", "Date"), where = "Close < 100"
  ) |>
  vg_params(
    brush = list(select = "intersect"), bandwidth = 0, pixelSize = 2,
    schemeColor = "pubugn", scaleColor = "sqrt"
  ) |>
  vg_vconcat(
    vg_hconcat(
      vg_slider(label = "Bandwidth (σ)", as = bandwidth, min = 0, max = 10, step = 0.1),
      vg_menu(label = "Pixel Size", as = pixel_size, options = c(0.5, 1, 2))
    ),
    vg_vspace(10),
    top_plot, bottom_plot
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/line-density.html) exactly,
aside from `data.stocks_after_2006.file` (`vg_data_url()` vs. a
relative path) and object key order.
</details>
