---
title: "Density 1D"
subtitle: "Density plots for over 200,000 flights, using kernel density estimation with an adjustable bandwidth."
format: html
engine: knitr
---

Recreates [Density 1D](https://idl.uw.edu/mosaic/examples/density1d.html)
from the mosaic/vgplot example gallery. Binning is performed
in-database, subsequent smoothing in-browser. The distance density
uses a log-scaled domain. To change the amount of smoothing, use the
slider to set the kernel bandwidth.

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)
bandwidth <- param(bandwidth)

delay_plot <-
  vg_mark_density_y(
    data_from = "flights", filter_by = brush, x = ~delay, fill = "#888", fill_opacity = 0.5, bandwidth = bandwidth
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(y_axis = NULL, x_domain = "Fixed", width = 600, margin_left = 10, height = 200)

distance_plot <-
  vg_mark_density_y(
    data_from = "flights", filter_by = brush, x = ~distance, fill = "#888", fill_opacity = 0.5, bandwidth = bandwidth
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(y_axis = NULL, x_scale = "log", x_domain = "Fixed", width = 600, margin_left = 10, height = 200)

spec <-
  vg_create() |>
  vg_meta(
    title = "Density 1D",
    description = paste(
      "Density plots (`densityY` mark) for over 200,000 flights, created using kernel",
      "density estimation. Binning is performned in-database, subsequent smoothing",
      "in-browser. The distance density uses a log-scaled domain. To change the amount of",
      "smoothing, use the slider to set the kernel bandwidth."
    )
  ) |>
  vg_data(name = "flights", file = vg_example_url("flights-200k.parquet")) |>
  vg_params(brush = list(select = "crossfilter"), bandwidth = 20) |>
  vg_vconcat(
    vg_slider(label = "Bandwidth (σ)", as = bandwidth, min = 0.1, max = 100, step = 0.1),
    delay_plot, distance_plot
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/density1d.html) exactly,
aside from `data.flights.file` (`vg_example_url()` vs. a relative path)
and object key order.
</details>
