---
title: "ECDF Plot for athlete heights"
subtitle: "Selection via plot elements, sorting, and fixed scale domains."
format: html
engine: knitr
---


Click on the bars below to select sports to be shown in the empirical
cumulative distribution plot on the right.
Multiple sports can be selected by using shift or command while clicking.

```{r}
#| label: ecdf-plot
#| code-fold: true

library(vgplotr)
spec <-
  vg_create() |>
  vg_data(name = "athletes", file = vg_example_url("athletes.parquet")) |>
  vg_params(sports = vg_selection("union")) |>
  vg_hconcat(
    vg_mark_bar_x(
      sport ~ vg_count(),
      fill = ~sport,
      sort = list(y = "-x", color = "-x")
    ) |>
      vg_toggle_y(as = param(sports)) |> # click a bar; Shift-click for more
      vg_highlight(by = param(sports)) |> # dim the bars not selected
      vg_scale_color(scheme = "turbo", domain = "Fixed") |>
      vg_plot(
        width = 320,
        height = 480,
        y_label = NULL,
        margin_left = 120
      ) |>
      vg_guide_x(
        label = "number of athletes",
        label_anchor = "center",
        label_arrow = FALSE
      ) |>
      vg_mark_grid_x(),
    vg_mark_line(
      y = ~ vg_cume_dist(order_by = "height", partition_by = ~sport),
      x = ~ sql("height * 100"),
      fy = ~sex,
      stroke = ~sport,
      sort = list(
        color = list(value = "y", reduce = "count", order = "descending")
      ),
      curve = "step-after",
      filter_by = param(sports),
      title = ~sport,
      tip = TRUE
    ) |>
      vg_guide_x(
        label = "height (cm)",
        label_anchor = "center",
        label_arrow = FALSE
      ) |>
      vg_guide_y(
        label = "cumulative proportion",
        label_anchor = "center",
        label_arrow = FALSE
      ) |>
      vg_plot(height = 480, width = 400, color_domain = "Fixed") |>
      vg_mark_frame() |>
      vg_scale_y(domain = c(0, 1)) |>
      vg_scale_color(scheme = "turbo")
  )

spec |> vg_render(height = 480)
```

:::{.callout-note}
#### Things to note:

* `domain = "Fixed"` -- keep the scale domain fixed (using all the data) rather than
updating it based on the selection.

* `vg_cume_dist(..., partition_by = ~ sport)` 
  calculates the cumulative distribution (for each sport) in DuckDB.

* `sql("height * 100")` convert from meters to centimeters (in DuckDB).

* `sort = list(...)` is used to make sure the color schemes for the bars 
and lines match. In that way, the bar graph also serves as a legend for color.

:::

<details>
<summary>JSON specification</summary>

```{r}
spec |> to_json() |> cat()
```

</details>

<details>
<summary>YAML specification</summary>

```{r}
spec |> to_yaml() |> cat()
```
</details>
