---
title: "U.S. Unemployment"
subtitle: "A choropleth map of unemployment rates for U.S. counties."
format: html
engine: knitr
---

Recreates [U.S.
Unemployment](https://idl.uw.edu/mosaic/examples/unemployment.html)
from the mosaic/vgplot example gallery (adapted from an [Observable
Plot example](https://observablehq.com/@observablehq/plot-us-choropleth)).
Requires the DuckDB `spatial` extension.

`combined` is a derived table (a plain SQL join query, mosaic-spec's
bare-string `DataQuery`) joining the spatial `counties` geometry with
the `rates` unemployment data by county id.

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "U.S. Unemployment",
    description = "A choropleth map of unemployment rates for U.S. counties. Requires the DuckDB `spatial` extension.",
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-us-choropleth)."
  ) |>
  vg_data(
    name = "counties",
    type = "spatial",
    file = vg_data_url("us-counties-10m.json"),
    layer = "counties"
  ) |>
  vg_data(
    name = "rates",
    file = vg_data_url("us-county-unemployment.parquet")
  ) |>
  vg_data(
    name = "combined",
    query = "SELECT a.geom AS geom, b.rate AS rate FROM counties AS a, rates AS b WHERE a.id = b.id\n"
  ) |>
  vg_vconcat(
    vg_legend_color(for_plot = "county-map", label = "Unemployment (%)"),
    vg_mark_geo(
      data_from = "combined",
      fill = ~rate,
      title = ~ sql("concat(rate, '%')")
    ) |>
      vg_plot(
        name = "county-map",
        margin = 0,
        color_scale = "quantile",
        color_n = 9,
        color_scheme = "blues",
        projection_type = "albers-usa"
      )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/unemployment.html) exactly,
aside from `data.counties.file`/`data.rates.file` (`vg_data_url()` vs.
a relative path) and object key order.
</details>
