---
title: "Earthquakes Feed"
subtitle: "Earthquake data from the USGS live feed, mapped over world land masses."
format: html
engine: knitr
---

Recreates [Earthquakes
Feed](https://idl.uw.edu/mosaic/examples/earthquakes-feed.html) from
the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-live-earthquake-map)).
To show land masses, this example loads and parses TopoJSON data in
the database. Requires the DuckDB `spatial` extension.

Both data sources use `vg_data(type = "spatial", ...)` -- mosaic's
`DataSpatial` source type, which loads GeoJSON/TopoJSON (and other
spatial formats DuckDB's `spatial` extension understands) directly into
the database. `layer = "land"` picks out one layer from the TopoJSON
file.

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "Earthquakes Feed",
    description = paste(
      "Earthquake data from the USGS live feed. To show land masses, this example loads and",
      "parses TopoJSON data in the database. Requires the DuckDB `spatial` extension."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-live-earthquake-map)."
  ) |>
  vg_data(
    name = "feed",
    type = "spatial",
    file = vg_data_url("usgs-feed.geojson")
  ) |>
  vg_data(
    name = "world",
    type = "spatial",
    file = vg_data_url("countries-110m.json"),
    layer = "land"
  ) |>
  vg_mark_geo(data_from = "world", fill = "currentColor", fill_opacity = 0.2) |>
  vg_mark_sphere(stroke_width = 0.5) |>
  vg_mark_geo(
    data_from = "feed",
    r = ~ sql("POW(10, mag)"),
    stroke = "red",
    fill = "red",
    fill_opacity = 0.2,
    title = ~title,
    href = ~url,
    target = "_blank"
  ) |>
  vg_plot(margin = 2, projection_type = "equirectangular")

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/earthquakes-feed.html)
exactly, aside from `data.feed.file`/`data.world.file` (`vg_data_url()`
vs. a relative path) and object key order.
</details>
