---
title: "Aeromagnetic Survey"
subtitle: "A raster visualization of the 1955 Great Britain aeromagnetic survey."
format: html
engine: knitr
---

Recreates [Aeromagnetic
Survey](https://idl.uw.edu/mosaic/examples/aeromagnetic-survey.html) from
the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-igfr90-raster)).

A raster visualization of the 1955 [Great Britain aeromagnetic
survey](https://www.bgs.ac.uk/datasets/gb-aeromagnetic-survey/), which
measured the Earth's magnetic field by plane. Each sample recorded the
longitude and latitude alongside the strength of the
[IGRF](https://www.ncei.noaa.gov/products/international-geomagnetic-reference-field)
in [nanoteslas](https://en.wikipedia.org/wiki/Tesla_(unit)). Demonstrates
both raster interpolation and smoothing (blur) options -- a menu and a
slider bind two `param()`s that are passed directly as mark option values
(`interpolate =`/`bandwidth =`), not just as encodings.

```{r setup}
library(vgplotr)
```

```{r}
interp <- param(interp)
blur <- param(blur)

spec <-
  vg_create() |>
  vg_meta(
    title = "Aeromagnetic Survey",
    description = paste(
      "A raster visualization of the 1955",
      "[Great Britain aeromagnetic survey](https://www.bgs.ac.uk/datasets/gb-aeromagnetic-survey/),",
      "which measured the Earth's magnetic field by plane. Each sample recorded the longitude and",
      "latitude alongside the strength of the",
      "[IGRF](https://www.ncei.noaa.gov/products/international-geomagnetic-reference-field) in",
      "[nanoteslas](https://en.wikipedia.org/wiki/Tesla_(unit)). This example demonstrates both",
      "raster interpolation and smoothing (blur) options."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-igfr90-raster)."
  ) |>
  vg_data(name = "ca55", file = vg_example_url("ca55-south.parquet")) |>
  vg_params(interp = "random-walk", blur = 0) |>
  vg_vconcat(
    vg_hconcat(
      vg_menu(
        label = "Interpolation Method",
        options = c("none", "nearest", "barycentric", "random-walk"),
        as = interp
      ),
      vg_hspace("1em"),
      vg_slider(label = "Blur", min = 0, max = 100, as = blur)
    ),
    vg_vspace("1em"),
    vg_mark_raster(
      data_from = "ca55",
      x = ~LONGITUDE,
      y = ~LATITUDE,
      fill = ~ vg_max(MAG_IGRF90),
      interpolate = interp,
      bandwidth = blur
    ) |>
      vg_plot(color_scale = "diverging", color_domain = "Fixed")
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/aeromagnetic-survey.html)
exactly, aside from `data.ca55.file` (a full URL from `vg_example_url()`),
object key order, and (cosmetically) how the multi-line `meta.description`
is assembled in R.
</details>
