---
title: "Sortable Table"
subtitle: "A sortable, \"infinite scroll\" table view over a backing database table."
format: html
engine: knitr
---

Recreates [Sortable Table](https://idl.uw.edu/mosaic/examples/table.html)
from the mosaic/vgplot example gallery. Click column headers to sort, or
command-click to reset the order; data is queried as needed as the table
is sorted or scrolled.

`vg_table()` is a layout-level input (like `vg_menu()`/`vg_slider()`), so
it can't be piped directly onto a `vgspec` -- it's combined with the rest
of the layout via `vg_vconcat()`/`vg_hconcat()`, even when (as here) it's
the only thing in the layout.

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "Sortable Table",
    description = paste(
      'A sortable, "infinite scroll" `table` view over a backing database table.',
      "Click column headers to sort, or command-click to reset the order.",
      "Data is queried as needed as the table is sorted or scrolled."
    )
  ) |>
  vg_data(name = "flights", file = vg_example_url("flights-200k.parquet")) |>
  vg_vconcat(vg_table(from = "flights", height = 300))

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Mosaic's own JSON spec puts the `table` input directly at the top level
(`{"input": "table", "from": "flights", "height": 300}`, no `vconcat`
wrapper), since it's the only thing in the layout. vgplotr's serializer
always wraps `vg_vconcat()`'s children in a `"vconcat": [...]` array, even
for a single child. Both are valid, semantically identical mosaic-spec
JSON. Aside from that and `data.flights.file`
(`vg_example_url()` vs. a relative path), this matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/table.html).
</details>
