---
title: "Seattle Weather Pivot"
subtitle: "A DuckDB PIVOT query reshapes daily weather observations into a cross-tab, shown in a sortable table."
format: html
engine: knitr
---

Recreates [Seattle Weather
Pivot](https://idl.uw.edu/mosaic/examples/seattle-weather-pivot.html)
from the mosaic/vgplot example gallery. A DuckDB `PIVOT` query reshapes
Seattle's daily weather observations into a cross-tab: one row per
year, with a column counting the days of each weather type. The
pivoted result is shown in a sortable `table` view. Click a column
header to sort.

```{r}
#| code-fold: true

library(vgplotr)
spec <-
  vg_create() |>
  vg_meta(
    title = "Seattle Weather Pivot",
    description = paste(
      "A DuckDB `PIVOT` query reshapes Seattle's daily weather observations into a",
      "cross-tab: one row per year, with a column counting the days of each weather type.",
      "The pivoted result is shown in a sortable `table` view. Click a column header to sort."
    )
  ) |>
  vg_data(name = "seattle_weather", file = vg_example_url("seattle-weather.parquet")) |>
  vg_data(
    name = "weatherByYear",
    query = paste0(
      "PIVOT (SELECT *, year(date) AS year FROM seattle_weather) ON weather IN ",
      "('drizzle', 'fog', 'rain', 'snow', 'sun') USING count(*) GROUP BY year ORDER BY year\n"
    )
  ) |>
  vg_table(
    data_from = "weatherByYear",
    align = list(year = "left"),
    width = list(year = 80),
    height = 180
  )

vg_render(spec, height = 120)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
spec |> to_json() |> cat()
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/seattle-weather-pivot.html)
exactly, aside from `data.seattle_weather.file` (`vg_example_url()` vs. a
relative path) and object key order.
</details>

<details>
<summary>YAML specification</summary>

```{r}
spec |> to_yaml() |> cat()
```
</details>
