---
title: "Scatter Plot Matrix (SPLOM)"
subtitle: "A 4x4 grid of linked scatter plots over four penguin measurements."
format: html
engine: knitr
---

Recreates [Scatter Plot Matrix
(SPLOM)](https://idl.uw.edu/mosaic/examples/splom.html) from the
mosaic/vgplot example gallery. Select a region to highlight
corresponding points across all plots.

Mosaic's own example hand-writes all 16 plots explicitly; here they're
built with a small loop over the four measurement columns instead
(`vg_plot_defaults()` supplies the attributes shared by every plot,
with only the leftmost column and bottommost row needing per-cell
overrides for axis labels and margins).

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)
columns <- c("bill_length", "bill_depth", "flipper_length", "body_mass")
rows <- rev(columns)

cell <- function(x_col, y_col, is_first_col, is_last_row) {
  p <-
    vg_mark_frame(stroke = "#ccc") |>
    vg_mark_dot(
      data_from = "penguins", x = as.formula(paste0("~", x_col)), y = as.formula(paste0("~", y_col)),
      fill = ~species, r = 2
    ) |>
    vg_interval_xy(as = brush) |>
    vg_highlight(by = brush, opacity = 0.1)

  attrs <- list()
  if (is_first_col) attrs <- c(attrs, list(y_axis = "left", margin_left = 45, width = 185))
  if (is_last_row) attrs <- c(attrs, list(x_axis = "bottom", height = 175, margin_bottom = 35))
  do.call(vg_plot, c(list(p), attrs))
}

grid_rows <- lapply(seq_along(rows), function(i) {
  cells <- lapply(seq_along(columns), function(j) {
    cell(columns[j], rows[i], is_first_col = (j == 1), is_last_row = (i == length(rows)))
  })
  do.call(vg_hconcat, cells)
})

spec <-
  vg_create() |>
  vg_meta(
    title = "Scatter Plot Matrix (SPLOM)",
    description = paste(
      "A scatter plot matrix enables inspection of pairwise bivariate distributions. Do",
      "points cluster or separate in some dimensions but not others? Select a region to",
      "highlight corresponding points across all plots."
    )
  ) |>
  vg_data(name = "penguins", file = vg_example_url("penguins.parquet")) |>
  vg_params(brush = list(select = "single")) |>
  vg_plot_defaults(
    x_ticks = 3, y_ticks = 4, x_domain = "Fixed", y_domain = "Fixed", color_domain = "Fixed",
    margin_top = 5, margin_bottom = 10, margin_left = 10, margin_right = 5,
    x_axis = NULL, y_axis = NULL, x_label_anchor = "center", y_label_anchor = "center",
    x_tick_format = "s", y_tick_format = "s", width = 150, height = 150
  )
spec <- do.call(vg_vconcat, c(list(spec), grid_rows))

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/splom.html) exactly (verified
cell-by-cell, including the leftmost-column and bottom-row attribute
overrides), aside from `data.penguins.file` (`vg_example_url()` vs. a
relative path) and object key order.
</details>
