---
title: "Earthquakes Globe"
subtitle: "A rotatable globe of earthquake activity."
format: html
engine: knitr
---

Recreates [Earthquakes
Globe](https://idl.uw.edu/mosaic/examples/earthquakes-globe.html) from
the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-earthquake-globe)).
To show land masses, this example loads and parses TopoJSON data in
the database. Requires the DuckDB `spatial` extension.

The `rotate` param combines two other params (`longitude`/`latitude`,
each driven by its own slider) into one value -- mosaic's own
`rotate: [$longitude, $latitude]` pattern -- passed straight through as
`vg_params(rotate = list(lon, lat))`.

```{r setup}
library(vgplotr)
```

```{r}
lon <- param(longitude)
lat <- param(latitude)
rotate_p <- param(rotate)

spec <-
  vg_create() |>
  vg_meta(
    title = "Earthquakes Globe",
    description = paste(
      "A rotatable globe of earthquake activity. To show land masses, this example loads",
      "and parses TopoJSON data in the database. Requires the DuckDB `spatial` extension."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-earthquake-globe)."
  ) |>
  vg_data(name = "earthquakes", file = vg_example_url("earthquakes.parquet")) |>
  vg_data(
    name = "land",
    type = "spatial",
    file = vg_example_url("countries-110m.json"),
    layer = "land"
  ) |>
  vg_params(longitude = -180, latitude = -30, rotate = list(lon, lat)) |>
  vg_vconcat(
    vg_hconcat(
      vg_slider(label = "Longitude", as = lon, min = -180, max = 180, step = 1),
      vg_slider(label = "Latitude", as = lat, min = -90, max = 90, step = 1)
    ),
    vg_mark_geo(
      data_from = "land",
      geometry = ~ vg_geojson(geom),
      fill = "currentColor",
      fill_opacity = 0.2
    ) |>
      # vg_mark_sphere() |>
      vg_mark_dot(
        data_from = "earthquakes",
        x = ~longitude,
        y = ~latitude,
        r = ~ sql("POW(10, magnitude)"),
        stroke = "red",
        fill = "red",
        fill_opacity = 0.2
      ) |>
      vg_plot(
        margin = 10,
        style = "overflow: visible;",
        projection_type = "orthographic",
        projection_rotate = rotate_p
      )
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/earthquakes-globe.html)
exactly, aside from `data.earthquakes.file`/`data.land.file`
(`vg_example_url()` vs. a relative path) and object key order.
</details>
