---
title: "Observable Latency"
subtitle: "Web request latency on Observable.com, linking a route-by-latency heatmap to a bar chart of top routes."
format: html
engine: knitr
---

Recreates [Observable
Latency](https://idl.uw.edu/mosaic/examples/observable-latency.html)
from the mosaic/vgplot example gallery (adapted from an [Observable
Framework example](https://observablehq.com/framework/examples/api/)).
Each pixel in the heatmap shows the most common route (URL pattern) at
a given response latency within a time interval. Use the bar chart of
most-requested routes to filter the heatmap and isolate specific
patterns. Or, select a range in the heatmap to show the corresponding
most-requested routes.

The `raster` mark's own `width`/`height` (2016x500, its internal pixel
grid resolution) are distinct from the *plot's* `width`/`height`
(680x300, its rendered size) -- these are per-mark properties, kept
separate from the plot-level attributes of the same name.

```{r setup}
library(vgplotr)
```

```{r}
filter_p <- param(filter)
highlight_p <- param(highlight)

latency_url <- "https://pub-1da360b43ceb401c809f68ca37c7f8a4.r2.dev/data/observable-latency.parquet"

heatmap_plot <-
  vg_mark_frame(fill = "black") |>
  vg_mark_raster(
    data_from = "latency", filter_by = filter_p, x = ~time, y = ~latency,
    fill = ~ vg_argmax(route, count), fill_opacity = ~ vg_sum(count),
    width = 2016, height = 500, image_rendering = "pixelated"
  ) |>
  vg_interval_xy(as = filter_p) |>
  vg_plot(
    color_domain = "Fixed", color_scheme = "observable10",
    opacity_domain = c(0, 25), opacity_clamp = TRUE,
    y_scale = "log", y_label = "↑ Duration (ms)", y_domain = c(0.5, 10000), y_tick_format = "s",
    x_scale = "utc", x_label = NULL, x_domain = c(1706227200000, 1706832000000),
    width = 680, height = 300, margins = list(left = 35, top = 20, bottom = 30, right = 20)
  )

routes_plot <-
  vg_mark_bar_x(
    data_from = "latency", filter_by = filter_p, x = ~ vg_sum(count), y = ~route, fill = ~route,
    sort = list(y = "-x", limit = 15)
  ) |>
  vg_toggle_y(as = filter_p) |>
  vg_toggle_y(as = highlight_p) |>
  vg_highlight(by = highlight_p) |>
  vg_plot(
    color_domain = "Fixed", x_label = "Routes by Total Requests", x_tick_format = "s", y_label = NULL,
    width = 680, height = 300, margin_top = 5, margin_left = 220, margin_bottom = 35
  )

spec <-
  vg_create() |>
  vg_meta(
    title = "Observable Latency",
    description = paste(
      "Web request latency on Observable.com.\n",
      "Each pixel in the heatmap shows the most common route (URL pattern) at a given",
      "response latency within a time interval.\n",
      "Use the bar chart of most-requested routes to filter the heatmap and isolate specific",
      "patterns.\n",
      "Or, select a range in the heatmap to show the corresponding most-requested routes."
    ),
    credit = "Adapted from an [Observable Framework example](https://observablehq.com/framework/examples/api/)."
  ) |>
  vg_data(name = "latency", file = latency_url) |>
  vg_params(filter = list(select = "crossfilter")) |>
  vg_vconcat(heatmap_plot, routes_plot)

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/observable-latency.html)
exactly, aside from object key order.
</details>
