---
title: "Sortable Table"
subtitle: "A sortable, \"infinite scroll\" table view over a backing database table."
format: html
engine: knitr
---

Recreates [Sortable Table](https://idl.uw.edu/mosaic/examples/table.html)
from the mosaic/vgplot example gallery. Click column headers to sort, or
command-click to reset the order; data is queried as needed as the table
is sorted or scrolled.

`vg_table()` is a layout-level input (like `vg_menu()`/`vg_slider()`).
Piped onto a spec that has no layout yet, as here, it becomes the spec's
whole layout; to combine it with plots or other inputs, use
`vg_vconcat()`/`vg_hconcat()` instead.

```{r}
#| code-fold: true

library(vgplotr)
spec <-
  vg_create() |>
  vg_meta(
    title = "Sortable Table",
    description = paste(
      'A sortable, "infinite scroll" `table` view over a backing database table.',
      "Click column headers to sort, or command-click to reset the order.",
      "Data is queried as needed as the table is sorted or scrolled."
    )
  ) |>
  vg_data(name = "flights", file = vg_example_url("flights-200k.parquet")) |>
  vg_table(data_from = "flights", height = 300)

vg_render(spec, height = 300)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
spec |> to_json() |> cat()
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/table.html) exactly, aside from
`data.flights.file` (`vg_example_url()` vs. a relative path) and object key
order.
</details>

<details>
<summary>YAML specification</summary>

```{r}
spec |> to_yaml() |> cat()
```
</details>
