---
title: "Seattle Weather"
subtitle: "An interactive view of Seattle's weather, linking a scatter plot to a bar chart of weather types."
format: html
engine: knitr
---

Recreates [Seattle Weather](https://idl.uw.edu/mosaic/examples/weather.html)
from the mosaic/vgplot example gallery (based on a [Vega-Lite/Altair
example](https://vega.github.io/vega-lite/examples/interactive_seattle_weather.html)
by Jake Vanderplas). By dragging on the scatter plot, you can see the
proportion of days in that range that have sun, fog, drizzle, rain, or
snow.

The color legend is embedded directly in the scatter plot's own mark
list (`vg_legend_color(as = click)`, no `for_plot`) rather than
standing alone -- clicking a legend entry toggles that weather type.
`domain`/`colors` are plain params (not Selections) holding a shared
category order and palette, referenced from both plots'
`color_domain`/`color_range`.

```{r setup}
library(vgplotr)
```

```{r}
click <- param(click)
range_p <- param(range)
domain <- param(domain)
colors <- param(colors)

scatter <-
  vg_mark_dot(
    data_from = "weather", filter_by = click, x = ~ vg_date_month_day(date), y = ~temp_max,
    fill = ~weather, r = ~precipitation, fill_opacity = 0.7
  ) |>
  vg_interval_x(as = range_p, brush = list(fill = "none", stroke = "#888")) |>
  vg_highlight(by = range_p, fill = "#ccc", fill_opacity = 0.2) |>
  vg_legend_color(as = click, columns = 1) |>
  vg_plot(
    xy_domain = "Fixed", x_tick_format = "%b", color_domain = domain, color_range = colors,
    r_domain = "Fixed", r_range = c(2, 10), width = 680, height = 300
  )

bars <-
  vg_mark_bar_x(data_from = "weather", x = ~ vg_count(), y = ~weather, fill = "#ccc", fill_opacity = 0.2) |>
  vg_mark_bar_x(data_from = "weather", filter_by = range_p, x = ~ vg_count(), y = ~weather, fill = ~weather) |>
  vg_toggle_y(as = click) |>
  vg_highlight(by = click) |>
  vg_plot(
    x_domain = "Fixed", y_domain = domain, y_label = NULL,
    color_domain = domain, color_range = colors, width = 680
  )

spec <-
  vg_create() |>
  vg_meta(
    title = "Seattle Weather",
    description = paste(
      "An interactive view of Seattle's weather, including maximum temperature, amount of",
      "precipitation, and type of weather. By dragging on the scatter plot, you can see the",
      "proportion of days in that range that have sun, fog, drizzle, rain, or snow."
    ),
    credit = "Based on a [Vega-Lite/Altair example](https://vega.github.io/vega-lite/examples/interactive_seattle_weather.html) by Jake Vanderplas."
  ) |>
  vg_data(name = "weather", file = vg_example_url("seattle-weather.parquet")) |>
  vg_params(
    click = list(select = "single"),
    domain = c("sun", "fog", "drizzle", "rain", "snow"),
    colors = c("#e7ba52", "#a7a7a7", "#aec7e8", "#1f77b4", "#9467bd")
  ) |>
  vg_vconcat(vg_hconcat(scatter), bars)

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/weather.html) exactly, aside
from `data.weather.file` (`vg_example_url()` vs. a relative path) and
object key order.
</details>
