---
title: "Protein Design Explorer"
subtitle: "Explore synthesized protein designs generated via RFDiffusion."
format: html
engine: knitr
---

Recreates [Protein Design
Explorer](https://idl.uw.edu/mosaic/examples/protein-design.html) from
the mosaic/vgplot example gallery (adapted from a UW CSE 512 project by
Christina Savvides, Alexander Shida, Riti Biswas, and Nora
McNamara-Bordewick). Explore synthesized proteins generated via
RFDiffusion to identify promising designs and assess the effects of
process parameters.

The color legend itself is a filter control here (`vg_legend_color(as
= query)`): clicking a legend entry toggles that version in/out of the
shared `query` crossfilter, same as the menus above it.

```{r setup}
library(vgplotr)
```

```{r}
query <- param(query)
point <- param(point)
plddt_domain <- param(plddt_domain)
pae_domain <- param(pae_domain)
scheme <- param(scheme)

controls <- vg_hconcat(
  vg_menu(from = "proteins", column = "partial_t", label = "Partial t", as = query),
  vg_menu(from = "proteins", column = "noise", label = "Noise", as = query),
  vg_menu(from = "proteins", column = "gradient_decay_function", label = "Gradient Decay", as = query),
  vg_menu(from = "proteins", column = "gradient_scale", label = "Gradient Scale", as = query)
)

top_hist <-
  vg_mark_rect_y(
    data_from = "proteins", filter_by = query,
    x = ~ vg_bin(plddt_total, steps = 60), y = ~ vg_count(), z = ~version, fill = ~version,
    order = "z", reverse = TRUE, inset_left = 0.5, inset_right = 0.5
  ) |>
  vg_plot(
    width = 600, height = 55, x_axis = NULL, y_axis = NULL, x_domain = plddt_domain,
    color_domain = "Fixed", color_scheme = scheme,
    margin_left = 40, margin_right = 0, margin_top = 0, margin_bottom = 0
  )

legend_row <- vg_hconcat(top_hist, vg_hspace(5), vg_legend_color(for_plot = "scatter", columns = 1, as = query))

scatter <-
  vg_mark_frame(stroke = "#ccc") |>
  vg_mark_raster(
    data_from = "proteins", filter_by = query, x = ~plddt_total, y = ~pae_interaction, fill = ~version, pad = 0
  ) |>
  vg_interval_xy(as = query, brush = list(stroke = "currentColor", fill = "transparent")) |>
  vg_mark_dot(
    data_from = "proteins", filter_by = point, x = ~plddt_total, y = ~pae_interaction,
    fill = ~version, stroke = "currentColor", stroke_width = 0.5
  ) |>
  vg_plot(
    name = "scatter", opacity_domain = c(0, 2), opacity_clamp = TRUE,
    color_domain = "Fixed", color_scheme = scheme, x_domain = plddt_domain, y_domain = pae_domain,
    x_label_anchor = "center", y_label_anchor = "center",
    margin_top = 0, margin_left = 40, margin_right = 0, width = 600, height = 450
  )

right_hist <-
  vg_mark_rect_x(
    data_from = "proteins", filter_by = query,
    x = ~ vg_count(), y = ~ vg_bin(pae_interaction, steps = 60), z = ~version, fill = ~version,
    order = "z", reverse = TRUE, inset_top = 0.5, inset_bottom = 0.5
  ) |>
  vg_plot(
    width = 55, height = 450, x_axis = NULL, y_axis = NULL,
    margin_top = 0, margin_left = 0, margin_right = 0, y_domain = pae_domain,
    color_domain = "Fixed", color_scheme = scheme
  )

tbl <- vg_table(
  as = point, filter_by = query, from = "proteins",
  columns = c(
    "version", "pae_interaction", "plddt_total", "noise",
    "gradient_decay_function", "gradient_scale", "movement"
  ),
  width = 680, height = 215
)

spec <-
  vg_create() |>
  vg_meta(
    title = "Protein Design Explorer",
    description = paste(
      "Explore synthesized proteins generated via RFDiffusion.",
      '"Minibinders" are small proteins that bind to a specific protein target.',
      "When designing a minibinder, a researcher inputs the structure of the target protein",
      "and other parameters into the AI diffusion model. The dashboard below enables",
      "exploration of the results to identify promising protein designs and assess the",
      "effects of process parameters."
    ),
    credit = paste(
      "Adapted from a [UW CSE 512](https://courses.cs.washington.edu/courses/cse512/24sp/)",
      "project by Christina Savvides, Alexander Shida, Riti Biswas, and Nora",
      "McNamara-Bordewick. Data from the",
      "[UW Institute for Protein Design](https://www.ipd.uw.edu/)."
    )
  ) |>
  vg_data(name = "proteins", file = vg_example_url("protein-design.parquet")) |>
  vg_params(
    query = list(select = "crossfilter"),
    point = list(select = "intersect", empty = TRUE),
    plddt_domain = c(67, 94.5), pae_domain = c(5, 29), scheme = "observable10"
  ) |>
  vg_vconcat(controls, vg_vspace("1.5em"), legend_row, vg_hconcat(scatter, right_hist), vg_vspace("1em"), tbl)

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/protein-design.html) exactly
(the full multi-paragraph description was abbreviated above for
readability, so `meta.description` text differs, though its structure
is identical), aside from `data.proteins.file` (`vg_example_url()` vs. a
relative path) and object key order.
</details>
