---
title: "Flights Hexbin"
subtitle: "Hexagonal bins show flight density by departure time and arrival delay, with marginal histograms."
format: html
engine: knitr
---

Recreates [Flights Hexbin](https://idl.uw.edu/mosaic/examples/flights-hexbin.html)
from the mosaic/vgplot example gallery. Select regions in the marginal
histograms to filter the density display.

`scale`'s initial value is declared as mosaic-spec's explicit
`ParamValue` object shape (`{"value": "log"}`, i.e. `list(value =
"log")`) rather than a bare string -- functionally identical to
`scale = "log"`, just a different way mosaic-spec allows writing the
same thing.

```{r setup}
library(vgplotr)
```

```{r}
scale <- param(scale)
query <- param(query)

time_hist <-
  vg_mark_rect_y(data_from = "flights", x = ~ vg_bin(time), y = ~ vg_count(), fill = "steelblue", inset = 0.5) |>
  vg_interval_x(as = query) |>
  vg_plot(
    margins = list(left = 5, right = 5, top = 30, bottom = 0), x_domain = "Fixed",
    x_axis = "top", y_axis = NULL, x_label_anchor = "center", width = 605, height = 70
  )

hexbins <-
  vg_mark_hexbin(data_from = "flights", filter_by = query, x = ~time, y = ~delay, fill = ~ vg_count(), bin_width = 10) |>
  vg_mark_hexgrid(bin_width = 10) |>
  vg_plot(
    name = "hexbins", color_scheme = "ylgnbu", color_scale = scale,
    margins = list(left = 5, right = 0, top = 0, bottom = 5),
    x_axis = NULL, y_axis = NULL, xy_domain = "Fixed", width = 600, height = 455
  )

delay_hist <-
  vg_mark_rect_x(data_from = "flights", x = ~ vg_count(), y = ~ vg_bin(delay), fill = "steelblue", inset = 0.5) |>
  vg_interval_y(as = query) |>
  vg_plot(
    margins = list(left = 0, right = 50, top = 4, bottom = 5), y_domain = c(-60, 180),
    x_axis = NULL, y_axis = "right", y_label_anchor = "center", width = 80, height = 455
  )

spec <-
  vg_create() |>
  vg_meta(
    title = "Flights Hexbin",
    description = paste(
      "Hexagonal bins show the density of over 200,000 flights by departure time and arrival",
      "delay. Select regions in the marginal histograms to filter the density display."
    )
  ) |>
  vg_data(name = "flights", file = vg_example_url("flights-200k.parquet")) |>
  vg_params(scale = list(value = "log"), query = list(select = "intersect")) |>
  vg_vconcat(
    vg_hconcat(
      vg_menu(label = "Color Scale", as = scale, options = c("log", "linear", "sqrt")),
      vg_hspace(10),
      vg_legend_color(for_plot = "hexbins")
    ),
    vg_hconcat(time_hist, vg_hspace(80)),
    vg_hconcat(hexbins, delay_hist)
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/flights-hexbin.html) exactly,
aside from `data.flights.file` (`vg_example_url()` vs. a relative path)
and object key order.
</details>
