---
title: "Gaia Star Catalog"
subtitle: "A 5M row sample of the Gaia star catalog, linking a sky map to a Hertzsprung-Russell diagram."
format: html
engine: knitr
---

Recreates [Gaia Star
Catalog](https://idl.uw.edu/mosaic/examples/gaia.html) from the
mosaic/vgplot example gallery. A `raster` sky map reveals our Milky
Way galaxy. Select high parallax stars in the histogram to reveal a
Hertzsprung-Russell diagram in the plot of stellar color vs. magnitude
on the right.

`color_scale`/`y_scale` are set to a **param reference**
(`scale_type`), not a literal string -- a menu-free way for one
control point (here just a param, not wired to a visible input in this
example) to switch every linked plot's scale type at once.

```{r setup}
library(vgplotr)
```

```{r}
brush <- param(brush)
bandwidth <- param(bandwidth)
pixel_size <- param(pixelSize)
scale_type <- param(scaleType)

gaia_query <- paste0(
  "-- compute u and v with natural earth projection\n",
  "WITH prep AS (\n",
  "  SELECT\n",
  "    radians((-l + 540) % 360 - 180) AS lambda,\n",
  "    radians(b) AS phi,\n",
  "    asin(sqrt(3)/2 * sin(phi)) AS t,\n",
  "    t^2 AS t2,\n",
  "    t2^3 AS t6,\n",
  "    *\n",
  "  FROM 'https://pub-1da360b43ceb401c809f68ca37c7f8a4.r2.dev/data/gaia-5m.parquet'\n",
  "  WHERE parallax BETWEEN -5 AND 20 AND phot_g_mean_mag IS NOT NULL AND bp_rp IS NOT NULL\n",
  ")\n",
  "SELECT\n",
  "  (1.340264 * \"lambda\" * cos(t)) / (sqrt(3)/2 * (1.340264 + (-0.081106 * 3 * t2) + ",
  "(t6 * (0.000893 * 7 + 0.003796 * 9 * t2)))) AS u,\n",
  "  t * (1.340264 + (-0.081106 * t2) + (t6 * (0.000893 + 0.003796 * t2))) AS v,\n",
  "  * EXCLUDE('t', 't2', 't6')\n",
  "FROM prep\n"
)

sky_plot <-
  vg_mark_raster(
    data_from = "gaia", filter_by = brush, x = ~u, y = ~v, fill = "density",
    bandwidth = bandwidth, pixel_size = pixel_size
  ) |>
  vg_interval_xy(pixel_size = 2, as = brush) |>
  vg_plot(
    xy_domain = "Fixed", color_scale = scale_type, color_scheme = "viridis",
    width = 440, height = 250, margin_left = 25, margin_top = 20, margin_right = 1
  )

mag_hist <-
  vg_mark_rect_y(
    data_from = "gaia", filter_by = brush,
    x = ~ vg_bin(phot_g_mean_mag), y = ~ vg_count(), fill = "steelblue", inset = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", y_scale = scale_type, y_grid = TRUE, width = 220, height = 120, margin_left = 65)

parallax_hist <-
  vg_mark_rect_y(
    data_from = "gaia", filter_by = brush,
    x = ~ vg_bin(parallax), y = ~ vg_count(), fill = "steelblue", inset = 0.5
  ) |>
  vg_interval_x(as = brush) |>
  vg_plot(x_domain = "Fixed", y_scale = scale_type, y_grid = TRUE, width = 220, height = 120, margin_left = 65)

hr_plot <-
  vg_mark_raster(
    data_from = "gaia", filter_by = brush, x = ~bp_rp, y = ~phot_g_mean_mag, fill = "density",
    bandwidth = bandwidth, pixel_size = pixel_size
  ) |>
  vg_interval_xy(pixel_size = 2, as = brush) |>
  vg_plot(
    xy_domain = "Fixed", color_scale = scale_type, color_scheme = "viridis", y_reverse = TRUE,
    width = 230, height = 370, margin_left = 25, margin_top = 20, margin_right = 1
  )

spec <-
  vg_create() |>
  vg_meta(
    title = "Gaia Star Catalog",
    description = paste(
      "A 5M row sample of the 1.8B element Gaia star catalog.\n",
      "A `raster` sky map reveals our Milky Way galaxy. Select high parallax stars in the",
      "histogram to reveal a Hertzsprung-Russel diagram in the plot of stellar color vs.",
      "magnitude on the right."
    )
  ) |>
  vg_data(name = "gaia", query = gaia_query) |>
  vg_params(brush = list(select = "crossfilter"), bandwidth = 0, pixelSize = 2, scaleType = "sqrt") |>
  vg_hconcat(
    vg_vconcat(sky_plot, vg_hconcat(mag_hist, parallax_hist)),
    vg_hspace(10),
    hr_plot
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON spec](https://idl.uw.edu/mosaic/examples/gaia.html)
exactly, aside from object key order.
</details>
