---
title: "Olympic Athletes"
subtitle: "An interactive dashboard of athlete statistics, linking menus, a search box, a scatter plot, and a sortable table."
format: html
engine: knitr
---

Recreates [Olympic
Athletes](https://idl.uw.edu/mosaic/examples/athletes.html) from the
mosaic/vgplot example gallery. The menus and searchbox filter the
display and are automatically populated by backing data columns.

`query`'s Selection declares `include = category` -- mosaic-spec's
`{"select": "intersect", "include": "$category"}` shape, where one
Selection's clauses build on another's. This is the same recursive
param-value serialization used for "Earthquakes Globe"'s
`rotate: [$longitude, $latitude]`, just nested one level differently
(a single param reference inside a Selection object, rather than an
array of two).

```{r setup}
library(vgplotr)
```

```{r}
category <- param(category)
query <- param(query)
hover <- param(hover)

controls <- vg_hconcat(
  vg_menu(label = "Sport", as = category, from = "athletes", column = "sport"),
  vg_menu(label = "Sex", as = category, from = "athletes", column = "sex"),
  vg_search(
    label = "Name", filter_by = category, as = query,
    from = "athletes", column = "name", type = "contains"
  )
)

scatter <-
  vg_mark_dot(
    data_from = "athletes", filter_by = query, x = ~weight, y = ~height, fill = ~sex, r = 2, opacity = 0.1
  ) |>
  vg_mark_regression_y(data_from = "athletes", filter_by = query, x = ~weight, y = ~height, stroke = ~sex) |>
  vg_interval_xy(as = query, brush = list(fillOpacity = 0, stroke = "black")) |>
  vg_mark_dot(
    data_from = "athletes", filter_by = hover, x = ~weight, y = ~height, fill = ~sex,
    stroke = "currentColor", stroke_width = 1, r = 3
  ) |>
  vg_plot(
    xy_domain = "Fixed", color_domain = "Fixed",
    margins = list(left = 35, top = 20, right = 1), width = 570, height = 350
  )

tbl <- vg_table(
  from = "athletes", max_width = 570, height = 250, filter_by = query, as = hover,
  columns = c("name", "nationality", "sex", "height", "weight", "sport"),
  width = list(name = 180, nationality = 100, sex = 50, height = 50, weight = 50, sport = 100)
)

spec <-
  vg_create() |>
  vg_meta(
    title = "Olympic Athletes",
    description = paste(
      "An interactive dashboard of athlete statistics. The menus and searchbox filter the",
      "display and are automatically populated by backing data columns."
    )
  ) |>
  vg_data(name = "athletes", file = vg_data_url("athletes.parquet")) |>
  vg_params(
    category = list(select = "intersect"),
    query = list(select = "intersect", include = category),
    hover = list(select = "intersect", empty = TRUE)
  ) |>
  vg_hconcat(
    vg_vconcat(controls, vg_vspace(10), scatter, vg_vspace(5), tbl)
  )

vg_render(spec)
```

<details>
<summary>Comparing to mosaic's own JSON spec</summary>

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/athletes.html) exactly, aside
from `data.athletes.file` (`vg_data_url()` vs. a relative path) and
object key order.
</details>
