---
title: "Computed Params"
subtitle: "Params computed from other params and data, and updated when they change."
format: html
engine: knitr
---

Some values in an interactive graphic depend on others: a reference line
at the average of the selected points, a target some factor above it, a
setting that should reset when the selection changes. In Mosaic's
JavaScript API this means attaching event listeners to params. In vgplotr,
`vg_computed_param()` and `vg_on_change()` do the same thing from R.

Brush a range of horsepower in the plot below. The solid line is the
average mpg of the cars inside the brush (of all cars when nothing is
brushed); the dashed line is that average times the factor chosen with the
slider. Brushing a new range resets the factor to 1.

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "Computed Params",
    description = paste(
      "The average mpg of the brushed cars, and a target that is a chosen",
      "factor times that average, both recomputed as the brush and slider change."
    )
  ) |>
  vg_data(name = "cars", data = mtcars) |>
  vg_params(factor = 1, brush = list(select = "intersect")) |>
  # The mean mpg of the cars the brush selects. With `data_from`, the formula can
  # use the data source's columns; with `filter_by`, only the selected rows
  # count, and the value is recomputed whenever the brush changes.
  vg_computed_param(
    sel_mean = ~ mean(mpg),
    data_from = "cars",
    filter_by = param(brush)
  ) |>
  # `factor` and `sel_mean` are params, so this is recomputed when either
  # one changes -- including when sel_mean itself was just recomputed.
  vg_computed_param(target = ~ factor * sel_mean) |>
  # When the brush changes, reset the slider's factor to 1.
  vg_on_change(param(brush), vg_set_param(param(factor), 1)) |>
  vg_vconcat(
    vg_slider(
      label = "Factor",
      as = param(factor),
      min = 0.5,
      max = 1.5,
      step = 0.05
    ),
    vg_mark_rule_x(x = 0) |>
      vg_mark_axis_x(label_anchor = "center", label_arrow = FALSE) |>
      vg_mark_rule_y(y = 0) |>
      vg_mark_axis_y(label_anchor = "center", label_arrow = FALSE) |>
      vg_mark_dot(data_from = "cars", x = ~hp, y = ~mpg, fill = "steelblue") |>
      # An interactor brushes the mark just before it -- here the dots.
      vg_interval_x(as = param(brush)) |>
      vg_mark_rule_y(
        y = param(sel_mean),
        stroke = "firebrick",
        stroke_width = 2
      ) |>
      vg_mark_rule_y(
        y = param(target),
        stroke = "firebrick",
        stroke_dasharray = "6 4"
      ) |>
      vg_plot(
        width = 600,
        height = 350,
        y_domain = c(0, 50),
        x_label = "Horsepower",
        y_label = "Miles per gallon"
      )
  )

vg_render(spec)
```

## How it works

A computed param's formula is ordinary R, translated to SQL and run by
DuckDB in the browser:

* A bare name is a param if one of that name has been declared with
  `vg_params()` (or is another computed param) -- here `factor` and
  `sel_mean`. Any other name is a column of the `data_from` data source -- here
  `mpg`.
* Aggregates such as `mean()`, `max()` or `n()` summarize the `data_from` data,
  restricted to the rows `filter_by` selects.
* When a param and a column share a name, write `param(name)` or
  `.data$name` to say which one you mean.

`vg_on_change()` runs its actions whenever its trigger (a param or a
selection) changes; each action is a `vg_set_param()`, with a constant or a
formula for the new value. See `?vg_computed_param` and `?vg_on_change` for
details.

<details>
<summary>JSON specification</summary>

```{r}
spec |> to_json(suppress_data = TRUE) |> cat()
```

Computed params and change handlers aren't part of Mosaic's spec, so they
are written under a top-level `"vgplotr"` key, and each computed param is
declared in `"params"` with an initial value of `null`. Use
`to_json(spec, vgplotr_keys = FALSE)` for a spec to use with Mosaic's own
tools. (`suppress_data = TRUE` leaves out the `mtcars` rows.)
</details>

<details>
<summary>YAML specification</summary>

```{r}
spec |> to_yaml(suppress_data = TRUE) |> cat()
```
</details>
