---
title: "Density 2D"
subtitle: "A 2D density plot where circle size indicates point density, split by species."
format: html
engine: knitr
---

Recreates [Density 2D](https://idl.uw.edu/mosaic/examples/density2d.html)
from the mosaic/vgplot example gallery. The data is divided by fill
color into three sets of densities. To change the amount of smoothing,
use the slider to set the kernel bandwidth.

The `density` mark's own `width`/`height` (its internal bin grid
resolution, bound to the `bins` slider) are distinct from the *plot's*
`width`/`height` (its rendered size) -- confirms the mark/plot
attribute collision fix (see `NOTES.md`) also works when the mark's
own property is bound to a param, not just a literal value.

```{r setup}
library(vgplotr)
```

```{r}
bandwidth <- param(bandwidth)
bins <- param(bins)

spec <-
  vg_create() |>
  vg_meta(
    title = "Density 2D",
    description = paste(
      "A 2D `density` plot in which circle size indicates the point density. The data is",
      "divided by fill color into three sets of densities. To change the amount of",
      "smoothing, use the slider to set the kernel bandwidth."
    )
  ) |>
  vg_data(name = "penguins", file = vg_data_url("penguins.parquet")) |>
  vg_params(bandwidth = 20, bins = 20) |>
  vg_vconcat(
    vg_hconcat(
      vg_slider(label = "Bandwidth (σ)", as = bandwidth, min = 1, max = 100),
      vg_slider(label = "Bins", as = bins, min = 10, max = 60)
    ),
    vg_mark_density(
      data_from = "penguins", x = ~bill_length, y = ~bill_depth, r = ~density, fill = ~species, fill_opacity = 0.5,
      width = bins, height = bins, bandwidth = bandwidth
    ) |>
      vg_mark_dot(data_from = "penguins", x = ~bill_length, y = ~bill_depth, fill = "currentColor", r = 1) |>
      vg_plot(
        r_range = c(0, 16), x_axis = "bottom", x_label_anchor = "center",
        y_axis = "right", y_label_anchor = "center",
        margins = list(top = 5, bottom = 30, left = 5, right = 50), width = 700, height = 480
      )
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/density2d.html) exactly,
aside from `data.penguins.file` (`vg_data_url()` vs. a relative path)
and object key order.
:::
