---
title: "Density Groups"
subtitle: "Density plots of penguin bill depths, grouped by species, with adjustable normalization and stacking."
format: html
engine: knitr
---

Recreates [Density Groups](https://idl.uw.edu/mosaic/examples/density-groups.html)
from the mosaic/vgplot example gallery. The `normalize` parameter
supports different forms of comparison, controlling if an individual
density estimate is scaled by total point mass or normalized by the
sum or max of the point mass. The `stack` and `offset` parameters
control stacking of density areas.

`offset`'s initial value is `NULL` (mosaic-spec's `"offset": null`) --
declared with `vg_params(offset = NULL)`, which needed a bug fix (see
`NOTES.md`) since it previously silently dropped the param instead of
declaring it with a `null` value.

```{r setup}
library(vgplotr)
```

```{r}
bandwidth <- param(bandwidth)
normalize <- param(normalize)
stack <- param(stack)
offset <- param(offset)

spec <-
  vg_create() |>
  vg_meta(
    title = "Density Groups",
    description = paste(
      "Density plots of penguin bill depths, grouped by species. The normalize parameter",
      "supports different forms of comparison, controlling if an individual density estimate",
      "is scaled by total point mass or normalized by the sum or max of the point mass. The",
      "stack and offset parameters control stacking of density areas."
    )
  ) |>
  vg_data(name = "penguins", file = vg_data_url("penguins.parquet")) |>
  vg_params(bandwidth = 20, normalize = "none", stack = FALSE, offset = NULL) |>
  vg_vconcat(
    vg_hconcat(
      vg_menu(label = "Normalize", as = normalize, options = c("none", "sum", "max")),
      vg_menu(label = "Stack", as = stack, options = c(FALSE, TRUE)),
      vg_menu(
        label = "Offset", as = offset,
        options = list(
          list(label = "none", value = NULL),
          list(label = "normalize", value = "normalize"),
          list(label = "center", value = "center")
        )
      )
    ),
    vg_mark_density_y(
      data_from = "penguins", x = ~bill_depth, fill = ~species, fill_opacity = 0.4,
      bandwidth = bandwidth, normalize = normalize, stack = stack, offset = offset
    ) |>
      vg_plot(margin_left = 50, height = 200)
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/density-groups.html) exactly,
aside from `data.penguins.file` (`vg_data_url()` vs. a relative path)
and object key order.
:::
