---
title: "Athlete Birth Waffle"
subtitle: "Waffle chart counting Olympic athletes based on which half-decade they were born."
format: html
engine: knitr
---

Recreates [Athlete Birth
Waffle](https://idl.uw.edu/mosaic/examples/athlete-birth-waffle.html) from
the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-waffle-unit)).

The inputs enable adjustment of waffle mark design options (`unit =`,
`round =`, `gap =`, `rx =`), each bound directly to a `param()`, and the
x-axis grouping uses a raw `sql()` expression (`year()` isn't one of
vgplotr's own transform functions).

```{r setup}
library(vgplotr)
```

```{r}
unit <- param(unit)
round <- param(round)
gap <- param(gap)
radius <- param(radius)

spec <-
  vg_create() |>
  vg_meta(
    title = "Athlete Birth Waffle",
    description = paste(
      "Waffle chart counting Olympic athletes based on which half-decade",
      "they were born. The inputs enable adjustment of waffle mark design options."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-waffle-unit)."
  ) |>
  vg_data(name = "athletes", file = vg_data_url("athletes.parquet")) |>
  vg_params(unit = 10, round = FALSE, gap = 1, radius = 0) |>
  vg_vconcat(
    vg_hconcat(
      vg_menu(label = "Unit", options = c(1, 2, 5, 10, 25, 50, 100), as = unit),
      vg_menu(label = "Round", options = c(TRUE, FALSE), as = round),
      vg_menu(label = "Gap", options = c(0, 1, 2, 3, 4, 5), as = gap),
      vg_slider(label = "Radius", min = 0, max = 10, step = 0.1, as = radius)
    ),
    vg_vspace(10),
    vg_mark_waffle_y(
      data_from = "athletes",
      unit = unit,
      round = round,
      gap = gap,
      rx = radius,
      x = ~ sql(r"(5 * floor(year("date_of_birth") / 5))"),
      y = ~ vg_count()
    ) |>
      vg_plot(xLabel = NULL, xTickSize = 0, xTickFormat = "d")
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/athlete-birth-waffle.html)
exactly, aside from `data.athletes.file` (`vg_data_url()` vs. a relative
path) and object key order.
:::
