---
title: "Axes & Gridlines"
subtitle: "Customized axis and gridline marks can be used in addition to standard scale attributes such as xAxis, yGrid, etc. Just add data!"
format: html
engine: knitr
---

Recreates [Axes &
Gridlines](https://idl.uw.edu/mosaic/examples/axes.html) from the
mosaic/vgplot example gallery. Demonstrates `axisX`/`axisY`/`gridX`/`gridY`
as standalone marks (for full control over placement/styling), and a
`ruleY` mark fed a literal inline data array (`data_from = c(0)`, mosaic's
`"data": [0]` shorthand) rather than a named data source.

```{r setup}
library(vgplotr)
```

```{r}
spec <-
  vg_create() |>
  vg_meta(
    title = "Axes & Gridlines",
    description = "Customized axis and gridline marks can be used in addition to standard scale attributes such as `xAxis`, `yGrid`, etc. Just add data!"
  ) |>
  vg_mark_grid_y(stroke_dasharray = "0.75 2", stroke_opacity = 1) |>
  vg_mark_axis_y(
    anchor = "left",
    tick_size = 0,
    dx = 38,
    dy = -4,
    line_anchor = "bottom"
  ) |>
  vg_mark_axis_y(
    anchor = "right",
    tick_size = 0,
    tick_padding = 5,
    label = "y-axis",
    label_anchor = "center"
  ) |>
  vg_mark_axis_x(label = "x-axis", label_anchor = "center") |>
  vg_mark_grid_x() |>
  vg_mark_rule_y(data_from = c(0)) |>
  vg_plot(
    xDomain = c(0, 100),
    yDomain = c(0, 100),
    xInsetLeft = 36,
    marginLeft = 0,
    marginRight = 35,
    width = 680
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/axes.html) exactly, aside from
object key order.
:::
