---
title: "Mark Types"
subtitle: "A subset of supported mark types."
format: html
engine: knitr
---

Recreates [Mark Types](https://idl.uw.edu/mosaic/examples/mark-types.html)
from the mosaic/vgplot example gallery.

Row 1: `barY`, `lineY`, `text`, `tickY`, `areaY`.
Row 2: `regressionY`, `hexbin`, `contour`, `heatmap`, `denseLine`.

```{r setup}
library(vgplotr)
```

```{r}
md <- data.frame(
  i = 0:7,
  u = LETTERS[1:8],
  v = c(2, 8, 3, 7, 5, 4, 6, 1)
)

row1 <- vg_hconcat(
  vg_mark_bar_y(data_from = "md", x = ~u, y = ~v, fill = "steelblue"),
  vg_mark_line_y(
    data_from = "md",
    x = ~u,
    y = ~v,
    stroke = "steelblue",
    curve = "monotone-x",
    marker = "circle"
  ),
  vg_mark_text(data_from = "md", x = ~u, y = ~v, text = ~u, fill = "steelblue"),
  vg_mark_tick_y(data_from = "md", x = ~u, y = ~v, stroke = "steelblue"),
  vg_mark_area_y(data_from = "md", x = ~u, y = ~v, fill = "steelblue")
)

row2 <- vg_hconcat(
  vg_mark_dot(
    data_from = "md",
    x = ~i,
    y = ~v,
    fill = "currentColor",
    r = 1.5
  ) |>
    vg_mark_regression_y(
      data_from = "md",
      x = ~i,
      y = ~v,
      stroke = "steelblue"
    ) |>
    vg_plot(xDomain = c(-0.5, 7.5)),
  vg_mark_hexgrid(stroke = "#aaa", stroke_opacity = 0.5) |>
    vg_mark_hexbin(data_from = "md", x = ~i, y = ~v, fill = ~ vg_count()) |>
    vg_plot(colorScheme = "blues", xDomain = c(-1, 8)),
  vg_mark_contour(
    data_from = "md",
    x = ~i,
    y = ~v,
    stroke = "steelblue",
    bandwidth = 15
  ) |>
    vg_plot(xDomain = c(-1, 8)),
  vg_mark_heatmap(
    data_from = "md",
    x = ~i,
    y = ~v,
    fill = "density",
    bandwidth = 15
  ) |>
    vg_plot(colorScheme = "blues", xDomain = c(-1, 8)),
  vg_mark_dense_line(
    data_from = "md",
    x = ~i,
    y = ~v,
    fill = "density",
    bandwidth = 2,
    pixel_size = 1
  ) |>
    vg_plot(colorScheme = "blues", xDomain = c(-1, 8))
)

spec <-
  vg_create() |>
  vg_meta(
    title = "Mark Types",
    description = "A subset of supported mark types."
  ) |>
  vg_data(name = "md", data = md) |>
  vg_vconcat(row1, row2) |>
  vg_plot_defaults(
    xAxis = NULL,
    yAxis = NULL,
    margins = list(left = 5, top = 5, right = 5, bottom = 5),
    width = 160,
    height = 100,
    yDomain = c(0, 9)
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

This matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/mark-types.html) for this example,
aside from two cosmetic (non-functional) differences: mosaic's serializer
collapses a single-mark, no-extra-attributes plot fragment inside a
`vconcat`/`hconcat` down to a bare mark object (`{"mark": "barY", ...}`),
while vgplotr always wraps it in `{"plot": [...]}`; and object key order
differs (alphabetical vs. call-argument order). Both are valid mosaic-spec
JSON and should render identically.
:::
