---
title: "Presidential Opinion"
subtitle: "Opinion poll data on historical U.S. presidents. Image marks are used to show presidential pictures."
format: html
engine: knitr
---

Recreates [Presidential
Opinion](https://idl.uw.edu/mosaic/examples/presidential-opinion.html)
from the mosaic/vgplot example gallery (adapted from an [Observable Plot
example](https://observablehq.com/@observablehq/plot-image-medals)).

Demonstrates an `image` mark, and a menu whose `options` are
label/value pairs (`list(label = ..., value = ...)`) rather than a plain
vector -- the selected value feeds into a `sql()` expression as an
interpolated `param()`.

```{r setup}
library(vgplotr)
```

```{r}
sign <- param(sign)

spec <-
  vg_create() |>
  vg_meta(
    title = "Presidential Opinion",
    description = paste(
      "Opinion poll data on historical U.S. presidents. Image marks are used to show",
      "presidential pictures. The dropdown menu toggles the opinion metric shown."
    ),
    credit = "Adapted from an [Observable Plot example](https://observablehq.com/@observablehq/plot-image-medals)."
  ) |>
  vg_data(
    name = "presidents",
    file = vg_data_url("us-president-favorability.parquet")
  ) |>
  vg_params(sign = 1) |>
  vg_vconcat(
    vg_mark_rule_y(data_from = c(0)) |>
      vg_mark_image(
        data_from = "presidents",
        x = ~`First Inauguration Date`,
        y = ~ sql(
          "\"Very Favorable %\" + \"Somewhat Favorable %\" + ",
          sign,
          " * (\"Very Unfavorable %\" + \"Somewhat Unfavorable %\")"
        ),
        src = ~`Portrait URL`,
        r = 20,
        preserve_aspect_ratio = "xMidYMin slice",
        title = ~Name
      ) |>
      vg_plot(
        x_inset = 20,
        x_label = "First inauguration date →",
        y_inset_top = 4,
        y_grid = TRUE,
        y_label = "↑ Opinion (%)",
        y_tick_format = "+f"
      ),
    vg_menu(
      label = "Opinion Metric",
      options = list(
        list(label = "Any Opinion", value = 1),
        list(label = "Net Favorability", value = -1)
      ),
      as = sign
    )
  )

vg_render(spec)
```

::: {.callout-note collapse="true"}
## Comparing to mosaic's own JSON spec

```{r}
cat(to_json(spec))
```

Matches [mosaic's own JSON
spec](https://idl.uw.edu/mosaic/examples/presidential-opinion.html)
exactly, aside from `data.presidents.file` (`vg_data_url()` vs. a relative
path) and object key order.
:::
